Desmodium Volatiles in “Push-Pull” Agriculture and Protection Against the Fall Armyworm, Spodoptera frugiperda

Version 1

Abstract

Abstract

Push-pull systems for sustainable pest management of crop plants employ repellent stimuli from intercrops (“push”) to repel herbivores and attract their predators and parasitoids, and attracting stimuli from border plants (“pull”) to lead herbivorous insects out of the crop. The most widespread implementation, intercropping with the legume Desmodium, reduces herbivory damage from the invasive fall armyworm (FAW) Spodoptera frugiperda. However, the three publications to date investigating underlying mechanisms disagree whether the Desmodium intercrop emits bioactive volatiles that repel FAW. We aimed to resolve this controversy by measuring volatile emission from Desmodium intortum (greenleaf Desmodium) and D. incanum intercrops in push-pull fields, and assaying their effects on the behavior of the FAW in oviposition and wind tunnel choice bioassays. We detected many volatile substances in the headspaces of both Desmodium species, which exhibited similar profiles, including substances previously reported to repel lepidopteran herbivores. FAW moths preferred to oviposit on maize over Desmodium, but not on maize further from versus closer to Desmodium plants that were inaccessible to the moths, but sharing the air. In a wind tunnel bioassay, the moths tended to prefer maize volatiles over a combination of volatiles from maize and D. intortum, but not D. incanum. In conclusion, we found D. intortum and D. incanum to emit volatiles that have been shown to be bioactive, and FAW moths to prefer maize over either Desmodium species. Moreover, additional mechanisms are likely important for reducing FAW damage to maize under push-pull cultivation.

Introduction

The sustainable intensification of agriculture is essential to achieve Sustainable Development Goal (SDG) 2, zero hunger, and reduce hidden costs of meeting requirements of other SDGs such as 13, climate action (FAO et al., 2020). This is particularly important in the developing Global South where larger yield gaps are observed compared to the developed countries with more financial resources (David Tilman et al., 2011). Intercropping may offer sustainable solutions with potential for strong positive effects on pest and disease control as well as associated biodiversity (Beillouin et al., 2021).

The concept of “push-pull” mixed cropping was first reported in 1987 as a “stimulo-deterrence” strategy for the reduction of pest damage in cotton, combining a pest repellent intercrop (“push“) interspersed with the main crop, and an attractant border crop (“pull“) to trap pests at the field perimeter (Bruce et al., 1987). As most widely practiced, current push-pull systems benefit from the attractive and repellent stimuli of perennial inter- and border crops (referred to as companion crops), thereby reducing or eliminating the need for pesticides (Pickett et al., 2014). In Kenya, the first push-pull system, introduced in 1997, was designed to reduce yield losses caused by stemborer species and comprised a combination of the main crop maize (Zea mays L), the intercrop molasses grass (Melinis minutiflora), and the border crop Napier grass (Pennisetum purpureum) or Sudan grass (Sorghum vulgare sudanense) (Khan et al., 1997b).

Several volatile substances emitted from molasses grass repelled female stem borers and simultaneously attracted their parasitoid Cotesia sesamiae (Khan et al., 1997a). Following the initial success of this system, there was interest in replacing molasses grass with a leguminous repellent intercrop that could improve soil quality by fixing nitrogen, and provide high-quality fodder. The legumes silverleaf Desmodium (Desmodium uncinatum) and its congeneric (Desmodium intortum) were shown to repel ovipositing stem borers and furthermore to suppress parasitic Striga weeds. (Khan et al., 2000, 2002). The first Desmodium-intercropped version of push-pull used D. uncinatum in place of M. minutiflora together with P. purpureum. As a result of screening for more drought-tolerant companion plants, a version using D. intortum combined with the border crop Brachiaria cv Mulato II was developed as “climate-adapted push-pull“, which has been demonstrated to increase maize yield by a factor of 2.5 (Midega et al., 2015). Recently, a “third generation” push-pull system was evaluated using the intercrop D. incanum to replace D. intortum – which does not flower and set seed in many parts of tropical Africa and thus limits adoption and spread of push-pull by limiting seed supply for the intercrop – with Brachiaria brizantha cv Xaraes as border crop, which is more resistant to herbivorous mites that attack the mulato cultivar (Cheruiyot et al., 2021).

The invasive fall armyworm (FAW), Spodoptera frugiperda, originating from a maize-specialized strain from northern and central America, has become a major threat to African maize crops since 2016 (Goergen et al., 2016; Day et al., 2017; Hailu et al., 2021; Zhang et al., 2023). It spread rapidly through East Africa, with most of the farmers in Ethiopia and Kenya encountering the FAW after the long rains in the first half of 2017 (Kumela et al., 2018). The invasiveness of the moth is fueled by its relatively short life cycle of about four weeks and the capability of the adult females to lay hundreds of eggs (Sparks, 1979). Other traits favoring the FAW’s spread in Africa and beyond are its strong flight capacity, lack of diapause, reported survival in diverse habitats, rapid development of resistance to insecticides / viruses, and the polyphagous nature (Wan et al., 2021). Although its wide host range of at least 353 host species across 76 plant families (Montezano et al., 2018), the high financial and yield losses on the main crop maize due to the pest are particularly devastating (Eschen et al., 2021). The FAW is reported to outcompete resident stemborers (Hailu et al., 2021; Sokame et al., 2021; Mutyambai et al., 2022). The use of pesticides is a popular approach to control the FAW, posing risks to environmental and human health, which includes acute pesticide-related illnesses as many smallholder farmers do not use personal protective equipment while spraying (Tambo et al., 2020). Therefore, sole reliance on pesticides is not sufficient to manage FAW sustainably, and Integrated Pest Management (IPM) strategies, such as the promotion of natural enemies, are desirable (Nyamutukwa et al., 2022; Van den Berg and du Plessis, 2022). Climate-adapted and third-generation push-pull systems have been reported to reduce plant damage and yield loss caused by the FAW (Midega et al., 2018; Hailu et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021; Mutyambai et al., 2022).

A systematic review on the chemical ecology of push-pull systems by Lang et al., (2022)) found only thirty publications (seven primary sources, and 23 publications reporting on results from these primary sources) on the chemistry of push-pull and related mixed-cropping systems, from which 206 compounds were reported to be potentially associated with push and pull effects. Of these, 101 were categorized as plant volatiles and reported by studies with a focus on plant-insect interactions (as opposed to studies with a focus on Striga control). However, none of these publications reported volatiles sampled under field conditions. Two recent papers, both published after the literature review conducted by Lang et al., (2022)), reported additional potentially bioactive compounds from companion plants in maize-Desmodium-grass push-pull systems, and both publications supported the hypothesis that volatiles from the companion crop D. intortum repel the FAW and attract parasitic wasps (Sobhy et al., 2022; Peter et al., 2023). In contrast, a third recent publication by Erdei et al., (2024) detected low levels of volatiles from D. intortum plants and found no consistent preference of FAW for maize over either D. intortum or maize together with D. intortum, and concluded that the protective effects of the intercrop were more likely to result from physical trapping of FAW larvae.

Thus, further investigations are needed to understand the mechanisms by which the maize-Desmodium-grass push-pull system can protect maize from FAW. A summary of the studies to date on push-pull effects on the FAW can be found in Figure 1. To our knowledge, no data on plant volatiles from the more drought-resistant D. incanum have been published so far. Therefore, this paper focuses on the volatiles of D. intortum and D. incanum and the effects of exposure to the headspaces of these plants, versus direct exposure to the plants, on the invasive pest FAW. We collected volatiles from plants in farmers’ fields as well as in bioassay setups under semi-field con-ditions, and conducted oviposition and flight tunnel bioassays to assess FAW moth preferences.

Studies to date on push-pull effects on the fall armyworm (FAW) embedded in the mechanisms according to the current state of knowledge of the system. Volatiles (and potentially other traits) of the intercrop repel the herbivorous insect and additionally attract its parasitoids, while volatiles (and potentially other traits) of the border crop attract herbivores away from the main crop (Pickett et al., 2014; Khan et al., 2018; Eigenbrode et al., 2016). The pull effect of the border crop Napier grass, Pennisetum purpureum was observed in earlier systems, as it attracted stemborers (Khan et al., 1997b), but could not be confirmed for the FAW with the border crop Brachiaria cv Mulato II (Sobhy et al., 2022). For more detail on all experiments, see supplementary file 1.

Results

Desmodium Volatile Profiles

A total of 25 substances were measured in at least 2/3 of field-collected samples from each of the Desmodium species, of which 11 occurred in both species. These substances include (E)-β- ocimene, (Z)-3-hexen-1-ol, (Z)-3-hexen-1-ol, acetate, 1-octen-3-ol, 3-octanone, caryophyllene, (3E)-4,8-dimethyl-1,3,7-nonatriene (DMNT), germacrene d, (3E,7E)-4,8,12-Trimethyl-1,3,7,11-tridecatetraene (TMTT), and five other peaks that were not fully identifiable. The unidentified peaks were cate-gorized based on mass spectra and relative retention times as a benzene derivative, a naphthalene derivative, a monoterpene, and two sequiterpenes (for feature information, see Table 1. The sesquiterpene (E)-α-bergamotene only occurred in D. incanum, while another unidentified sesquiterpene only occurred in D. intortum (see Figure 2 and for a per-sample heatmap Figure 2—figure Supplement 1). Of these substances, (Z)-1,5-octadien-3-ol, 1,2,3-trimethylbenzene, 3-pentanol, and hexyl acetate are here for the first time reported in connection with the chemical ecology of pushpull cropping systems (Lang et al., 2022; Sobhy et al., 2022; Peter et al., 2023). Germacrene D and 1-hexanol were not yet reported from Desmodium species, but were reported in relation to mixed cropping for pest control. Germacrene D was detected in elevated levels in maize plants grown in soil from push-pull fields in comparison with those grown in soil from maize monocultures (Mutyambai et al., 2019), while germacrene D and 1-hexanol both were reported to be emitted from bean plants, Vicia faba, and used by the black bean aphid, Aphis fabae, for host detection (Pickett and Khan, 2016; Webster et al., 2008).

Heatmap comparing the log10-transformed peak area of the non-zero hits of substances present in at least 2/3 of the samples of at least one of the Desmodium species in the field, showing relative peak areas detected in field- or pot-grown (bioassay conditions) D. intortum and D. incanum, with pot-grown maize for comparison. For the field Desmodium samples all substances occurring in less then 2/3 of all samples were removed and here displayed in grey, while lower frequency was accepted for the potted samples and the controls that showed a smaller sampling size. The x-axis displays the species and their growth conditions (field or pot) and the y-axis shows the target substances in alphabetical order. The color indicates mean of the log-10 transformed peak area of all samples with the substance present. The grey color indicates that a substance occurs in more than one, but less than 2/3 of the samples from the field. “Maize Infested” refers to maize plants that were exposed to FAW eggs, as moths were allowed to oviposit on the plants two nights prior to volatile sampling. Sample sizes: Ambient Control Field = 4, D. incanum Field = 14, D. incanum Pot = 3, D. intortum Field = 11, D. intortum Pot = 5, Maize infested Pot = 4, Maize Pot = 4.

Heatmap comparing the log10-transformed peak area of all samples. The x-axis displays the species and their growth conditions (field or pot) and the y-axis shows the target substances in alphabetical order. The color indicates mean of the log-10 transformed peak area, whereby transparency indicates that no peak was found.

Oviposition Bioassays

Oviposition choice bioassays were conducted to determine the influence of direct or indirect contact with Desmodium plants on the oviposition behavior of FAW moths. A fraction of the potted Desmodium plants used in the oviposition assays was sampled for volatiles the night before the start of trials, and the composition of these samples was compared with the substances and features detected from Desmodium plants sampled in farmers’ fields. In the comparison between field Desmodium plants, and potted bioassay plants that were sampled in a smaller number, the majority of substances showed a higher relative abundance in potted D. incanum (bioassay conditions), whereas in D. intortum, some substances occurred in higher concentrations in the potted plants, while others were not detected. Sixteen of the volatiles detected in Desmodium plants in farmers’ fields were also present in either uninfested or infested maize plants, with 14 substances overlapping (see Figure 3).

A: Principal component analysis plot based on the normalized peak areas of the 25 target substances. Sample sizes: Ambient Control Field = 4, D. incanum Field = 14, D. incanum Pot = 3, D. intortum Field = 11, D. intortum Pot = 5, Maize infested Pot = 4, Maize Pot = 4 B: Loading plot of the projection of the variables on the first two dimensions. The x-axis displays PC1 (Dimension 1) and the y-axis shows PC2 (Dimension 2). The color indicates the cos2, whereby dark and long arrows indicate a better representation of a loading on these first two dimensions.

The count of eggs and egg batches showed that fewer eggs were laid directly on Desmodium plants, with the egg count on maize plants being 7.9 or 6.8 times higher than on D. incanum or D. intortum, respectively (see Figure 4 and Figure 4—figure Supplement 1). A mixed model with egg counts as the observed variable and accounting for plant position detected a significant difference between the control and all other treatments (F(1) = 9.21, p = 0.003), as well as for the comparison of the indirect and the direct treatments (F(1) = 9.32, p = 0.003). However, no significant difference was found between the two Desmodium species (F(1) = 1.01, p = 0.32) (see Table 3).

A: Setup of the control with two maize plants inside the cage. B: Setup of the direct treatments with one maize and one Desmodium plant inside the cage. C: Setup of the indirect treatments with two maize plants placed inside the cage and one Desmodium plant placed in proximity to one of the maize plants, but unreachable for the moths. D: The boxplots display the relative number of eggs (light grey boxes) or egg batches (dark grey boxes) per position (x axis) and treatment (superordinate boxes). The lower and upper hinges correspond to the 25th and 75th percentiles, while the whiskers extend to the largest and smallest non-outlier values, respectively. Outliers are values outside a window of 1.5 x the interquartile range and are plotted individually. Desmodium is abbreviated with ‘Des’ and treatment is abbreviated with ‘treat’. See Figure 4—figure Supplement 1 for a breakdown of eggs per repetition and batch. Sample sizes (n, replicate units of a given treatment): Control = 21, D. incanum direct = 19, D. intortum direct = 20, D. incanum indirect = 19, D. intortum indirect = 20.

A: Variance of total egg count per repetition. The x-axis shows the sum of all egg found in all position within one repetition. The grey points display the total egg count per repetition and the black diamonds show the mean value over all data points per treatment. B: Boxplot including individual data points of the egg batch size (egg number per batch) per location (x-axis) and per treatment (superordinate boxes). Desmodium is abbreviated with ‘Des’ and treatments is abbreviated with ‘treat’. Samples sizes: Control = 21, D. incanum direct = 19, D. intortum direct = 20, D. incanum indirect = 19, D. intortum indirect = 20.

Upper Line: Schematic drawing of the placement of the plants in the cages according to the treatments. The rectangle represents the cage, the plant in the beige pot represents maize and the plant in the red pot Desmodium. Lower Line: Photos of the experimental setup of three treatments control (left), D. incanum direct (center) and D. intortum indirect (right)

To minimize volatile interference of neighboring setups, the treatments were placed in groups next to each other (represented by letters) with the distance of at least one cage length in between, which corresponded to 60 – 100 cm. Due to space limitations on two tables the cages of group H were placed slightly closer with an approximate distance of 50 cm. The positions of the treatments were swapped between each cycle so that treatment was conducted at each position once. In the treatments where Desmodium plants were placed outside the cages, the Desmodium plants were normally placed between the cages and in greenhouse 1B additionally on the outside of the outermost cage of a group (as can be seen on the cages of G). However, as all treatments were photographed, in retrospect several misplaced plants were noticed (see cage K2). Therefore, it must be considered that only conclusions about the influence of Desmodium volatiles in proximity versus those in further distance of >0.5 m can be made and any volatiles reaching longer distances might have affected neighboring replications. Greenhouse 1B: Schematic graph of the cage setup in the greenhouse 1B (above) and a photograph taken of the cages H2, J2, J3, K1 & K2 (right). This greenhouse provided space for 15 cages in total, whereby three replicates of each treatment could be repeated per cycle. The plants display the placement of the plants in the replicates of the second cycle from 23. – 26.05.2023. Lightgrey squares = tables, darkgrey squares = cages, yellow circle = maize plants, green circle = Desmodium plants; The red circle indicated the situation where a plant was misplaced and therefore was in the closest distance of the neighboring repetition.

To minimize volatile interference of neighboring setups, the treatments were placed in groups next to each other (represented by letters) with the distance of at least one cage length in between, which corresponded to 60 – 100 cm. Due to space limitations on two tables the cages of group E were placed slightly closer with an approximate distance of 50 cm. The positions of the treatments were swapped between each cycle, with all treatments being conducted in 4/5 positions. In the treatments where Desmodium plants were placed outside the cages, the Desmodium plants were normally placed between the cages and in greenhouse 1B additionally on the outside of the outermost cage of a group (as can be seen on the cages of G). However, as all treatments were photographed, in retrospect several misplaced plants were noticed (see Figure 4—figure Supplement 3 Figure of Greenhouse 1B, cage K2). Therefore, it must be considered that only conclusions about the influence of Desmodium volatiles in proximity versus those in further distance of >0.5 m can be made and any volatiles reaching longer distances might have affected neighboring replications. Greenhouse 3B: Schematic graph of the cage setup in the greenhouse 3B (left) and a photograph taken from the door toward the cages A1 & A2 (right). This greenhouse consisted of 10 cages in total, whereby two replicates of each treatment could be repeated per cycle. The plants display the placement of the plants in the cycle of 23. – 26.05.2023. lightgrey squares = tables, darkgrey squares = cages, yellow circle = maize plants; green circle = Desmodium plants.

Fall armyworm eggs are normally laid staggered in clusters and with hair-like scales on the surface texture which complicates accurate egg counting (see picture A). Therefore, eggs were collected with sticky tape to separate the layers and spread all eggs out to one dimension (B). The egg batches were taped to white paper, taking care not to squash eggs and photographed in a UV imaging system (Syngene, Cambridge, England) against UV light coming from underneath the paper (C). A script for semiautomatic counts in ImageJ (Version 1.54f, National Institutes of Health, USA) was developed and used for counting the exact number of eggs per picture (D).

Wind Tunnel Bioassays

Wind tunnel bioassays were conducted to determine short-term effects of plant volatiles on the flight behavior of FAW moths.

Two-thirds of the moths tested showed ten or fewer segment changes during the five minutes of each experiment. Across all treatments, at least 77% of the moths showed no movement in the last 2 min of the experiment or only little activity, with a maximum of two segment changes in the last 3 min of the experiment, which was interpreted to mean that they had made a decision within the five-minute timeframe of the experiment. Overall, moths showed a reduction in segment changes over time, particularly for the comparison of maize vs. maize + D. intortum (see Figure 5). Across all treatments, a few individuals displayed high flight activity of up to 76 segment changes with no indication of decline over time.

A: Line plot displaying the number of segment changes per minute until the experiment ended after 5 min. The line width indicates the number of overlaying lines and the colors indicate the moth activity level: ‘High Activity’ = Moths showed higher activity with at least 6 segment changes in the last 3 min of the experiment. ‘Little Activity’ = Moths showed max. 2 segment changes in the last 3 min of the experiment. ‘Settled after 3 min’ = No segment changes after 3 min. ‘Settled after 2 min’ = No segment changes after 2 min. The mean of all repetitions is represented by the black dashed line. B: Setup of the wind tunnel for a Desmodium treatment comparing a maize plant (left) vs. a maize and a Desmodium plant (right). The moths are released through an opening in the center (at 0 cm) and observed for 5 min. C: Violin plot displaying the duration of stays in each segment in treatments with D. incanum and D. intortum (comparison of maize and maize + Desmodium) with the inclusion of the individual data points per repetition (black) and their mean (red): The x-axis displays the location in the wind tunnel and the y-axis the total duration that each moth stayed in each segment. Below the x-axis, the last position of the moths that were considered settled (settled after <3min or little activity) is displayed. See Figure 5—figure Supplement 1 for the set-up and segmentation of the two-choice wind tunnel setup. Sample sizes (n, replicate units of a given treatment: Control = 18, D. incanum = 18, D. intortum = 19.)

Experimental setup of the two-choice wind tunnel with a D. incanum treatment on the left and a maize on the right. The photograph was taken in the light, while the experiments were conducted in the dark only with red light. Yellow marks: Air exhaust at the centre of the wind tunnel covered with a fine-meshed net. Red marks: Two red light bulbs placed symmetrical approx. 20 cm above the upper rim of the wind tunnel. White marks: Wind tunnel body (30 x 30 x 100 cm) and its imaginary separation of the length in five sectors of 20 cm. Green marks: Air transfer from the top of the plant bags via teflon tubing to the wind tunnel. Blue marks: Air inlet for charcoal-filtered air provided by Volatile Collection Kits via teflon tubing.

No significant effects were detected when comparing the stays in the maize segments versus the Desmodium segments across the treatments using a mixed model (see Table 4). However, a preference for the side with maize is apparent for the comparison of maize vs. maize + D.intortum, especially when comparing the mean stay duration as well as the final settling location of the moth, whereas moths showed a slight preference for maize + D. incanum instead of maize alone (see Figure 5). Fourteen of nineteen moths exposed to airflow from maize vs. maize + D.intortum settled in a segment of the wind tunnel within the first three minutes of the experiment, a larger proportion than in the maize vs. maize control or the maize vs. maize + D. incanum treatment.

Discussion

The fall armyworm (FAW) Spodoptera frugiperda invaded sub-Saharan Africa in 2016 and is responsible for major yield losses in maize crops (Day et al., 2017; Goergen et al., 2016). Push-pull cropping of maize with Desmodium and border grasses is reported to reduce FAW damage (Midega et al., 2015; Cheruiyot et al., 2021; Mutyambai et al., 2022; Yeboah et al., 2021; Hailu et al., 2018) and to increase maize yield (Midega et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021). Push-pull cropping systems benefit from the attractive and repellent stimuli of perennial companion crops (Pickett et al., 2014), with more than one hundred plant volatiles from different classes, especially terpenoids and fatty acid-derived “green leaf” volatiles, reported to be associated with push-pull effects (Lang et al., 2022). Nevertheless, the repellent role of volatiles from greenleaf Desmodium (D. intortum) has been investigated in only three studies and almost exclusively in potted plants, with conflicting reports of whether these volatiles repel FAW moths, and the only study reporting field measurements recovering few volatiles in relatively low abundance (Sobhy et al., 2022; Peter et al., 2023; Erdei et al., 2024). Here, we characterized volatiles sampled from both D. intortum (greenleaf Desmodium) and D. incanum growing in farmer fields or in pots, and conducted bioassays with FAW moths to determine whether the presence of Desmodium plants or their volatiles would alter moth preferences for maize. We identified up to 17 - 19 substances in headspace samples from Desmodium plants, often in abundance similar to or greater than volatiles measured from maize plants under comparable conditions, and we observed that FAW moths preferred to oviposit on maize rather than on Desmodium plants of either species. However, proximity to a Desmodium plant sharing the airspace did not influence FAW oviposition preferences for maize in a greenhouse cage choice assay, nor did FAW moths show a significant preference for maize volatiles over volatiles from a combination of maize and Desmodium plants in a wind tunnel choice assay. We note that these assays were performed in a risk area for malaria and other mosquito-borne illnesses, and so experimenters were protected with DEET, which has low volatility but is active in the vapor phase, where it can cause host avoidance by binding to insect olfactory receptors (although this has been mostly studied in the dipterans Aedes aegypti and Drosophila melanogaster) (DeGennaro, 2015). We reduced any potential influence on the lepidopteran in our experiments by using minimal amounts of DEET, avoiding direct contact of any DEET with experimental materials, and conducting behavioral assays without the insects being exposed to volatiles from the experimenter (experimenters were generally not present during oviposition assays, and the insects were in an isolated airspace during wind tunnel bioassays). We conclude that Desmodium plants in push-pull cultivation emit known bioactive volatiles, but cannot confirm that these volatiles alone repel FAW moths based on our bioassays, and therefore suggest that other features of the maize-Desmodium-border grass push-pull system, including the influence of border grasses as well as belowground interactions and other repellent traits of Desmodium, may be important for protecting maize from FAW moths and their larvae.

Emission of Potentially Bioactive Volatiles by Desmodium Intercrops

The headspace of Desmodium plants was dynamically sampled on Tenax TA adsorbent directly in push-pull fields, as well as from potted plants used for oviposition bioassays, overnight.

All substances occurring in fewer than 2/3 of the field samples were excluded from further analysis, which represents an arbitrary limit. Quantitative comparisons within single substances are possible, but must be handled with care, as normalization based on the biomass and leaf area of the plants was not possible. The chosen adsorbent, Tenax TA, is known to be effective for lipophilic to medium-polarity organic compounds of intermediate molecular weight (ca. C7-C26) (Dettmer and Engewald, 2002; Tholl et al., 2020). It is possible that headspace samples changed during storage (up to 3 months tightly closed at room temperature) as degradation or thermal rearrangements on Tenax TA material have been reported before (Dettmer and Engewald, 2002; Alborn et al., 2021). We extracted headspace samples using thermal desorption, which is simple and sensitive, but not appropriate for some compounds, such as the sesquiterpene germacrene A, which are susceptible to thermal rearrangement or degradation (Tholl et al., 2006).

Despite these caveats, our samples yielded a variety of volatiles with high signal:noise ratios, of which 13 had been previously reported from sampling the headspace of potted D. intortum plants onto Porapak Q for sampling durations of 24 h or 48 h followed by solvent extraction and GC-MS analysis (Sobhy et al., 2022; Peter et al., 2023). In contrast, Erdei et al., (2024) detected fewer volatiles in lower relative abundance from D. intortum using solid-phase microextraction (SPME) and a saturation time of 18 - 24h to sample intact, mechanically damaged or herbivoryinduced potted plants or intact plants in Tanzanian and Ugandan fields. Our study is the only one of these that limited the window of sampling strictly to the nighttime hours corresponding to the reported activity window for the FAW(Sparks, 1979). In pretests no shorter activity window could be determined as the moths still showed mating behaviour and potential oviposition at around midnight (see Appendix 1). Differences might be explained due to the different use of sampling techniques, as the sensitivity of SPME is lower in comparison to dynamic headspace collection (Tholl et al., 2006). To our knowledge, the volatile profile of D. incanum was not previously reported. In conclusion, the hypothesis that D. intortum and D. incanum emit potentially bioactive volatiles is supported.

Oviposition Choice Bioassay

FAW oviposition behaviour was observed over three nights in two treatments per Desmodium species. In one treatment, FAW moths had direct contact with Desmodium plants and in another, they had only indirect contact through aerial exposure to plants outside of, but directly next to a mesh cage. Moths always had direct contact with maize plants.

Critical consideration must be given to differences in the volatile profiles between the potted plants and those measured in farmers’ fields, which led to a separation of potted and field D. intortum in the second dimension of a PCA, and a near-separation of D. incanum in the first dimension. As the untargeted analysis was conducted based on the field Desmodium samples, any volatile substances that only occur in the potted plants were not identified or analyzed further (full results of GC-MS analysis are available with the source data for this publication). Oviposition on maize can alter the maize volatile profile and affect volatile emission in the following days, which is why the maize samples are divided into plants before and after (infested) oviposition exposure. Of the substances occurring in field Desmodium plants, approximately half were abundant in healthy and infested potted maize plants, with only one substance, (E)-α-Bergamotene, being present solely in infested maize plants. The sampling of potted plants was conducted using the same procedure as in the field, but in a greenhouse with netted walls on two sides that allowed air circulation as for the oviposition bioassays. Overall, this highlights the importance of studying oviposition cues and moth preferences directly in push-pull fields in the future.

In all treatments, the number of eggs laid per batch showed no difference depending on the position, which indicates that FAW moths did not adjust their behavior once they started laying eggs. This differs slightly from the outcomes reported by Peter et al., (2023), where a reduction in egg number per batch was observed and the number of eggs, but not the number of egg batches, differed between maize vs. maize + D. intortum. However, a clear preference for maize compared to the two Desmodium species was observed in the treatments where FAW moths had direct contact with both plant species, with a large effect size of 7 – 8 times more eggs on maize than on Desmodium. The preference of maize over D. intortum in oviposition assays with direct contact to both plants is consistent with various publications using similarly controlled setups (Sobhy et al., 2022; Peter et al., 2023; Erdei et al., 2024). There was little oviposition on the cage in our bioassays, which indicates that moths primarily chose between plants in our setup, but differs from the strong preference found by Sobhy et al., (2022) of FAW moths to oviposit directly on the mesh of a bioassay cage rather than on the plants when offered maize co-planted with D. intortum. Desmodium and maize plants were planted in the same pots by Sobhy et al., (2022), and therefore maize volatiles may have been affected by belowground interactions or through priming by Desmodium volatiles prior to or during oviposition bioassays. This is in fact closer to the configuration in push-pull fields, but does not isolate the influence of the Desmodium headspace, as we sought to do here.

In our bioassays, no significant effect could be seen in the treatments where moths had only indirect contact with Desmodium volatiles coming from a Desmodium plant proximate to one maize plant, but outside of the cage. This is not consistent with the hypothesis that Desmodium volatiles repel FAW moths as proposed (Pickett et al., 2014; Sobhy et al., 2022). It is possible that the distance of about 60 cm between the plants in the cages was not suitable to create a gradient of volatiles sufficient to allow moths to choose between maize plants closer to versus further from Desmodium, in which case neighboring treatments with distances of 50 – 100 cm could have also affected each other. To our knowledge, the only comparable setup was reported by Erdei et al., (2024), in which two maize plants were exposed to volatiles of either D. intortum or an artificial plant in a modified wind tunnel, and moths did not show a significant preference for either maize plant.

In conclusion, we found a preference of FAW for maize over D. intortum and D. incanum, but this preference could not be solely attributed to Desmodium volatiles in our experiments. It is not to be expected that herbivores use all potential host plants equally, and preferences among host plants or for host over non-host plants can be based on different plant traits. FAW moths may also prefer maize over Desmodium due to other traits such as the surface texture, as proposed in the findings from Erdei et al., (2024) that silicate trichomes on D. intortum pose a danger to FAW larvae; or chemical repellents directly on the leaf surface. Belowground interactions or priming between Desmodium and maize plants co-planted in push-pull fields may also change moths’ preference for maize, an effect which could not have been observed with our experimental design, but might explain inconsistencies between our findings and those from Erdei et al., (2024) versus those from Sobhy et al., (2022).

Wind Tunnel Bioassay

The short-term flight behavior of FAW moths was tested in a two-choice wind tunnel setup where volatiles of a maize plant vs. maize + Desmodium were compared against each other. Flight behavior showed large variation among the individual moths, but three-quarters of the moths had settled after at least three minutes or showed fewer than three segment changes in the last two minutes of the experiment. This experimental setup has limited space and duration, and moths may behave differently over longer times or when given more space to maneuver, as under normal field conditions. Furthermore, the quarter of the moths that showed high activity indicate the difficulty of quantifying the responses of such active fliers. Interestingly, a larger proportion of moths settled after 2 minutes when presented with maize versus maize + D. intortum, rather than continuing to fly. The moths showed a weak preference for the volatiles of the single maize plant over the volatiles of maize + D. intortum, whereby they slightly preferred volatiles from D. incanum + maize over maize alone but did not distinguish between volatiles from two maize plants (negative control). However, none of the treatments showed any significant preference. In contrast, repulsion of moths by D. intortum volatiles with or without the addition of maize was observed in no-choice assays when the total distance flown upwind at the end of the experiment was used as a metric (Sobhy et al., 2022; Peter et al., 2023). Based on these wind tunnel results, it cannot be excluded that the exchange of D. intortum with D. incanum in third-generation push-pull fields may reduce protection against FAW moths. However, in direct oviposition choice assays (in which moths were enclosed in a cage with one maize and one Desmodium plant), maize was preferred equally over either Desmodium species.

Conclusion

Several field studies have described positive effects of push-pull systems with the companion crops D. intortum or D. incanum on maize yield (Midega et al., 2015, 2018; Cheruiyot et al., 2021). Here, we show that 17 - 19 volatile substances that have been previously shown to affect pest host choice were found in the headspaces of both companion crops D. intortum and D. incanum within the activity window of FAW moths. However, we did not observe a clear repellent effect of Desmodium volatiles on the FAW in our bioassays. FAW moths clearly preferred to oviposit on maize over D. intortum or D. incanum plants. Our results thus indicate that moths prefer maize over Desmodium and that this may be influenced by short-range mechanisms such as an unfavorable leaf surface of Desmodium. However, the main interest in the context of push-pull is how Desmodium contributes to protecting co-planted maize. A reduction of FAW damage on maize under push-pull cultivation has been demonstrated in multiple studies (Midega et al., 2018; Hailu et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021; Mutyambai et al., 2022). Based on the literature, the pull effect of border crop volatiles may be important for reducing FAW damage in push-pull fields, in combination with the attraction of parasitoids by Desmodium volatiles and unfavorable host qualities of Desmodium plants for FAW (Figure 1). Furthermore, the Desmodium intercrop may be equally or more important for weed suppression, soil fertility, and possibly improving maize plant health and rigor as for volatile-mediated repellence of FAW in the push-pull system.

Methods and materials

All volatile sampling and bioassays were performed with two Desmodium species, Desmodium intortum (greenleaf Desmodium) and Desmodium incanum (often referred to as African Desmodium by the push-pull farmers) and in the method section referred to as Desmodium for simplicity. All raw data and code used for the statistical analysis can be found on Zenodo (CERN, Geneva, Switzerland, https://doi.org/10.5281/zenodo.11633890) and Github (github.com/ DariaMOdermatt/DesmodiumVolatilesinPush-Pull).

Pretest Moth Activity Window

The determination of the activity window of the FAW was relevant for the choice of the duration of the volatile sampling. The FAW is reported to be nocturnal (Sparks, 1979), without report of a more precise activity window. For the first cycle, two single maize leaves of the landrace ‘Jowi white’ were placed in wooden cages with meshed side walls (100 x 60 x 60 cm) in high cylindrical glasses filled with water, but sealed with a piece of cotton to prevent moths of entering the glasses. Moths of the age of 3 - 4 days were released in pairs of one female and one male in 10 different cages (resulting in 20 moths in total) and observed from 7pm - 11.30pm in two consecutive nights. The moths were recollected in falcon tubes in the morning after the first night, sealed with a piece of cotton to allow air circulation and stored at a temperature of approx. 25 °C until they were released in the same combinations on the following evening. For the second night fresh maize leaves were provided for the moths to oviposit. For the second cycle, potted maize plants with multiple leaves of the landrace ‘Jowi white’ were placed inside the wooden cages with a piece of cotton soaked with water. Three female and two male moths of different ages were placed each in three cages (resulting in 15 moths in total) and observed between 6pm - 1am in two consecutive nights. During day time the moths and the plant were left unaltered in the cage.

Volatile Sampling

Volatiles of Desmodium were collected from plants in push-pull fields of local smallholder farmers near Mbita in the counties Homa bay and Siaya (coordinates determined with WGS84 coordinate system (latitude, longitude): D. incanum = (-0.382096, 34.175487), (-0.4298279, 34.207891), (0.189957, 34.36072), D. intortum = (-0.382096, 34.175487), (-0.6302736, 34.494595), (-0.551479, 34.314673)). Desmodium volatiles were collected on three different push-pull fields in May and June 2023 during the long rains, selecting four representative plants per field. As it is becoming more common, some of the push-pull fields were mixed with vegetables such as kale or cowpeas. Volatile collections were also performed from potted Desmodium plants, as well as from healthy and infested potted maize plants on the research campus. Five three-week-old maize plants, later labeled as“infested”, were exposed in two cages, each with each three female and two male moths, for two nights prior to the volatile sampling, which resulted in 1 to 10 batches per plant. Potted plants were placed in a greenhouse with a glass roof and two glass walls, in which air was able to circulate through netted walls on the opposing longer walls of the greenhouse.

As an adsorbent, commercially available Tenax TA glass tubes (containing 150 mg adsorbent, 35 – 60 mesh, Markes International Ltd, England or 186 mg adsorbent, 60 – 80 mesh, Merck, Germany) were used. A handful of leaves were enclosed in a preheated roasting bag (Sainsbury’s, London, UK). The inlet tubing pushed 500 mL/min charcoal-filtered air into the bags from the lower bag rim, while the Tenax TA tube was connected to an air outlet tubing through a small hole at the top of the bag. Air flows were regulated by PYE Volatile Collection Kits (B. J. Pye, Kings Walden, UK) as described in Steen et al. (2019). Volatiles were accumulated on the adsorbent by aspirating the air at 200 mL/min (+/- 40 mL/min) for 11.75 h (+/- 30 min) overnight from 7 pm until 6.45 am of the following day. Under the same conditions, four ambient controls were sampled with an empty roasting bag on two fields with each Desmodium species. Additionally, a Tenax TA storage control was stored with all tubes, but never taken to the fields.

TD-GC-MS Measurement

All samples were measured on a TD-GC-MS instrument (Thermal Desorption – Gas Chromatography – Mass Spectrometer) from Shimadzu (TD30-QP2020NX Kyoto, Japan). The Tenax TA tubes were desorbed for 15 min at 220 °C under 80 mL/min nitrogen flow. Volatiles were trapped on a Tenax cooling trap at -20 °C and after desorption was completed, the cooling trap was heated rapidly to 230 °C. The sample was injected with a split ratio of 1:5 and separated on a Rtx-Wax column (30 m, 0.25 mmID, 0.25 μm, Restek Corporation, USA) with the following 35 min oven program: holding at 40 °C for 2 min, heating to 150 °C with 10 °C/min, holding at 150 °C for 2 min, heating to 190 °C with 3 °C/min, heating to 230 °C with 10 °C/min and holding at 230 °C for 3 min. Mass fragments from 33 m/z to 400 m/z resulting from electron ionization at 70 eV were recorded with a scan rate of 5 Hz at an ion source temperature of 230 °C from 3 – 30 min.

Feature Detection

A feature list was generated in MZmine (Version 3.9.0, (Schmid et al., 2023)) from all compounds that occurred in at least one-third of the field samples of either Desmodium species and in less than 4/5 of the combined control samples (ambient control and storage control). Features were numbered and considered as unidentified if there was no corresponding reference compound. After removal of contaminants, the features were added to an existing target list that was composed with commercially available plant volatiles that can be mainly classified as terpenoids and benzenoids, which as a whole contain a large variety of functional groups such as alcohols, aldehydes, ketones and esters. Additionally, the existing list included four green leaf volatiles ((E)-2-hexen-1-ol acetate, (E)-2-hexenal, (Z)-3-hexen-1-ol, (Z)-3-hexen-1-ol acetate) and a selection of more specific substances reported in push-pull fields such as (E)-4,8–dimethyl–1,3,7-nonatriene (DMNT) and (E,E)-4,8,12-trimethyl-1,3,7,11-tridecatetraene (TMTT). Integration of all substances and features was manually checked for the all control samples (ambient and storage control) using the LabSolutions Insight GCMS software (Shimadzu corporation, Kyoto, Japan). Peaks for all targets that occurred in at least 1/2 of the Desmodium field samples and in no more than 3/5 of controls were manually checked for all samples. Eventually, all targets that occured in at least 2/3 of the field samples of D. intortum or D. incanum and in no more than 3/5 of the controls were considered as present. In Table 1 the electron impact ionization (EI)-MS data of the unidentified features are displayed, and in Table 2 the origin of all reference standards used to identify the target hits can be found.

Oviposition Bioassays

Plants

All plants were planted in plastic pots in black cotton soil in Mbita, Kenya. Maize plants (SC Duma 43, Seed Co Limited, Nairobi, Kenya) were grown from seeds with the addition of fertilizer (3 g DAP 18-46-0, Yara East Africa LTD, Nairobi, Kenya) and used at the age of 2 – 4 weeks with five to seven fully-grown leaves. Desmodium plants were obtained from a push-pull field on the icipe Mbita campus and kept in pots without fertilizer.

Fall Armyworm Moths

To form the FAW colony, wild individuals were collected in the counties Siaya, Kisumu, Migori, and Vihiga in Western Kenya. Larvae were fed on artificial diet based on soy flour, wheat germ, raw linseed oil, mineral mix, sugar, aureomycin, vitamins, agar, methyl parabene, sorbic acid, and calcium propionate (Article No 870-265-3747, Southland Products Inc, Lake Village AR, USA). Moths were fed on water and in rare cases after emergence on 10% honey solution. The colony was occasionally restocked with wild FAW moths. Each day after hatching, moths of both sexes were transferred to a cage to allow mating until the start of the experiment.

Experimental Procedure

Cages (100 x 60 x 60 cm) lined with wooden floors and ceilings and netted walls were placed in greenhouses that allowed air circulation through mesh side walls (for pictures of the real setup, see Figure 4—figure Supplement 2, for exact positioning in the greenhouses, see Figure 4—figure Supplement 3 and Figure 4—figure Supplement 4). Two plants were placed within the cage at the greatest distance possible (approx. 60 cm) in two opposite corners. For the control treatment, two maize plants were placed in the cage, while the direct Desmodium treatment consisted of one maize plant and one Desmodium plant in the cage. In two additional treatments with indirect contact, two maize plants were placed in the cage, while one Desmodium plant was placed outside the cage unreachable for the FAW moths, but in proximity to one maize plant. Three female and two male moths at the age of 4 – 5 days were released approx. 1 h before dusk and allowed to oviposit under natural conditions of L12:D12 for three consecutive nights. Eggs were collected in five different groups depending on the position in which they were laid, either on one of the two plants, close to one of the plants within a maximum of 20 cm distance, or on the cage further than 20 cm from any plant (labeled as “No Decision”). Each treatment was repeated 19 – 21 times in total over five cycles, with up to 5 replicates per treatment per cycle.

Egg Count

Each batch of FAW eggs of was collected with sticky tape to separate the layers and spread all eggs out to one dimension. Each egg batch was taped to white paper and photographed in a UV imaging system (Syngene, Cambridge, England) against UV light coming from underneath the paper. A script for semiautomatic counts in ImageJ (Version 1.54f, National Institutes of Health, USA) was developed and used for counting the exact number of eggs per picture (see Figure 4—figure Supplement 5). The code is available in the data depsitory.

Wind tunnel Bioassays

Plants

All plants were planted from seeds without the addition of fertilizer in plastic pots in standard red soil mixed with manure (ratio of 2:1) at the icipe campus in Nairobi, Kenya. Maize plants (SC Duma 43, Seed Co Limited, Nairobi, Kenya) were used at the age of 4 – 7 weeks with five to eight fullygrown leaves. It was observed that maize plants seemed to grow more slowly in Nairobi than in Mbita, which could be explained by the generally lower temperatures in Nairobi. D. intortum and D. incanum seeds were obtained from a push-pull field on the icipe campus Mbita and were 6 – 8 weeks and 7 months old when used, respectively.

Fall Armyworm Moths

To form the FAW colony, wild individuals were collected in Central Kenya such as in the counties Kiambu, Muranga, and Embu and restocked with wild individuals every three months. Larvae were reared on maize leaves, while moths were only fed with water. The colony was occasionally restocked with wild FAW moths. Each day after hatching, moths of both sexes were transferred to a cage to allow mating until the start of the experiment. At least 2 h prior to the experiment, female moths were separated from males and and placed in the darkened experimental room for adjustment to temperature and humidity.

Experimental Procedure

The five upper leaves of a maize plant were wrapped in a preheated roasting bag (Sainsbury’s, London, UK) with the addition of approximately the same biomass of Desmodium on one side of a two-choice wind tunnel (100 x 30 x 30 cm) and compared to a single maize plant wrapped in the same manner on the other side. Control treatments were conducted with one maize plant wrapped in this way on each side. Air was pushed through activated charcoal filters and into the roasting bag from the lower rim of the bag at a rate of 740 – 820 mL/min and a Teflon tube conveyed the plant volatiles from the top of the roasting bag to one side of the wind tunnel. Inside the wind tunnel, the air was pumped out from the center at the 0 cm mark resulting in an air stream at the air transfer tubing of 480 – 570 mL/min with no greater difference than 4 mL/min between the two sides. One female moth at an age of 4 – 5 days was released through a hole in the center at 0 cm and its flight behavior was observed for 5 minutes. The wind tunnel was separated into 5 sectors of 20 cm each, and a time stamp was set every time the moth changed the segment (See Figure 5—figure Supplement 1). All experiments were conducted between 7 pm and 1:30 am under red light. Each trial was recorded on video using a mobile phone camera (Fairphone 4, 48 MP), which captured the wind tunnel from a front view. The videos were rewatched for the data acquisition and in cases where moths were barely visible, complemented with the notes from the live observation. To simplify the data analysis, the two left sectors (50 – 30 cm left & 30 – 10 cm left) were combined, as were the two right sectors (10 – 30 cm right, 30 – 50 cm right).

Statistical analysis

All statistical analysis was performed in RStudio (R version 4.3.2, RStudio version 2024.04.2). Mixed models were performed to determine significant effects of the positions of FAW eggs in the oviposition bioassays or for the duration of stay in the different sectors in the wind tunnel bioassays. In both bioassays a control with two maize plants was performed. Therefore, the control replications were included twice in the dataset, once with the egg count data or the duration of stay of both sides interchanged. In addition, the replication was included as a random effect into the model equations. For the oviposition bioassay, the egg counts on the maize or the Desmodium plant were compared amongst all treatments, with the inclusion of the greenhouse, the cage (combination of the cage group and the cage number), the start date and the replication number as random effects. For the wind tunnel bioassay, the duration of stay closer to the maize or the Desmodium plants were compared amongst all treatments, with the inclusion of the start date and the replication number as random effects.

eLife Assessment

Research on push-pull systems often focuses on controlled environments, limiting our understanding of their effectiveness under real-world conditions. This important study has validated how push-pull systems work in natural settings. However, the manuscript remains incomplete, since the findings have only been partially supported, as acknowledged by the authors.

Reviewer #1 (Public review):

Summary:

The manuscript of Odermatt et al. investigates the volatiles released by two species of Desmodium plants and the response of herbivores to maize plants alone or in combination with these species. The results show that Desmodium releases volatiles in both the laboratory and the field. Maize grown in the laboratory also released volatiles, in a similar range. While female moths preferred to oviposit on maize, the authors found no evidence that Desmodium volatiles played a role in lowering attraction to or oviposition on maize.

Strengths:

The manuscript is a response to recently published papers that presented conflicting results with respect to whether Desmodium releases volatiles constitutively or in response to biotic stress, the level at which such volatiles are released, and the behavioral effect it has on the fall armyworm. These questions are relevant as Desmodium is used in a textbook example of pest-suppressive sustainable intercropping technology called push-pull, which has supported tens of thousands of smallholder farmers in suppressing moth pests in maize. A large number of research papers over more than two decades have implied that Desmodium suppresses herbivores in push-pull intercropping through the release of large amounts of volatiles that repel herbivores. This premise has been questioned in recent papers. Odermatt et al. thus contribute to this discussion by testing the role of odors in oviposition choice. The paper confirms that ovipositing FAW preferred maize, and also confirmed that odors released from Desmodium appeared not important in their bioassays.

The paper is a welcome addition to the literature and adds quality headspace analyses of Desmodium from the laboratory and the field. Furthermore, the authors, some of whom have since long contributed to developing push-pull, also find that Desmodium odors are not significant in their choice between maize plants. This advances our knowledge of the mechanisms through which push-pull suppresses herbivores, which is critically important to evolving the technique to fit different farming systems and translating this mechanism to fit with other crops and in other geographical areas.

Weaknesses:

Below I outline the major concerns:

(1) Clear induction of the experimental plants, and lack of reflective discussion around this: from literature data and previous studies of maize and Desmodium, it is clear that the plants used in this study, particularly the Desmodium, were induced. Maize appeared to be primarily manually damaged, possibly due to sampling (release of GLV, but little to no terpenoids, which is indicative of mostly physical stress and damage, for example, one of the coauthor's own paper Tamiru et al. 2011), whereas Desmodium releases a blend of many compounds (many terpenoids indicative of herbivore induction). Erdei et al. also clearly show that under controlled conditions maize, silver leaf and green leaf Desmodium release volatiles in very low amounts. While the condition of the plants in Odermatt et al. may be reflective of situations in push-pull fields, the authors should elaborate on the above in the discussion (see comments) such that the readers understand that the plant's condition during the experiments. This is particularly important because it has been assumed that Desmodium releases typical herbivore-induced volatiles constitutively, which is not the case (see Erdei et al. 2024). This reflection is currently lacking in the manuscript.

(2) Lack of controls that would have provided context to the data: The experiments lack important controls that would have helped in the interpretation:

(2a) The authors did not control the conditions of the plants. To understand the release of volatiles and their importance in the field, the authors should have included controlled herbivory in both maize and Desmodium. This would have placed the current volatile profiles in a herbivory context. Now the volatile measurements hang in midair, leading to discussions that are not well anchored (and should be rephrased thoroughly, see eg lines 183-188). It is well known that maize releases only very low levels of volatiles without abiotic and biotic stressors. However, this changes upon stress (GLVs by direct, physical damage and eg terpenoids upon herbivory, see above). Erdei et al. confirm this pattern in Desmodium. Not having these controls, means that the authors need to put the data in the context of what has been published (see above).

(2b) It would also have been better if the authors had sampled maize from the field while sampling Desmodium. Together with the above point (inclusion of herbivore-induced maize and Desmodium), the levels of volatile release by Desmodium would have been placed into context.

(2c) To put the volatiles release in the context of push-pull, it would have been important to sample other plants which are frequently used as intercrop by smallholder farmers, but which are not considered effective as push crops, particularly edible legumes. Sampling the headspace of these plants, both 'clean' and herbivore-induced, would have provided a context to the volatiles that Desmodium (induced) releases in the field - one would expect unsuccessful push crops to not release any of these 'bioactive' volatiles (although 'bioactive' should be avoided) if these odors are responsible for the pest suppressive effect of Desmodium. Many edible intercrops have been tested to increase the adoption of push-pull technology but with little success.

Because of the lack of the above, the conclusions the authors can draw from their data are weakened. The data are still valuable in the current discussion around push-pull, provided that a proper context is given in the discussion along the points above.

(3) 'Tendency' of the authors to accept the odor hypothesis (i.e. that Desmodium odors are responsible for repelling FAW and thereby reduce infestation in maize under push-pull management) in spite of their own data: The authors tested the effects of odor in oviposition choice, both in a cage assay and in a 'wind tunnel'. From the cage experiments, it is clear that FAW preferred maize over Desmodium, confirming other reports (including Erdei et al. 2024). However, when choosing between two maize plants, one of which was placed next to Desmodium to which FAW has no tactile (taste, structure, etc), FAW chose equally. Similarly in their wind tunnel setup (this term should not be used to describe the assay, see below), no preference was found either between maize odor in the presence or absence of Desmodium. This too confirms results obtained by Erdei et al. (but add an important element to it by using Desmodium plants that had been induced and released volatiles, contrary to Erdei et al. 2024). Even though no support was found for repellency by Desmodium odors, the authors in many instances in the manuscript (lines 30-33, 164-169, 202, 279, 284, 304-307, 311-312, 320) appear to elevate non-significant tendencies as being important. This is misleading readers into thinking that these interactions were significant and in fact confirming this in the discussion. The authors should stay true to their own data obtained when testing the hypothesis of whether odors play a role in the pest-suppressive effect of push-pull.

(4) Oviposition bioassay: with so many assays in close proximity, it is hard to certify that the experiments are independent. Please discuss this in the appropriate place in the discussion.

(5) The wind tunnel has a number of issues (besides being poorly detailed):

(5a) The setup which the authors refer to as a 'wind tunnel' does not qualify as a wind tunnel. First, there is no directional flow: there are two flows entering the setup at opposite sides. Second, the flow is way too low for moths to orient in (in a wind tunnel wind should be presented as a directional cue. Only around 1.5 l/min enters the wind tunnel in a volume of 90 l approximately, which does not create any directional flow. Solution: change 'wind tunnel' throughout the text to a dual choice setup /assay.

(5b) There is no control over the flows in the flight section of the setup. It is very well possible that moths at the release point may only sense one of the 'options'. Please discuss this.

(5c) Too low a flow (1,5 l per minute) implies a largely stagnant air, which means cross-contamination between experiments. An experiment takes 5 minutes, but it takes minimally 1.5 hours at these flows to replace the flight chamber air (but in reality much longer as the fresh air does not replace the old air, but mixes with it). The setup does not seem to be equipped with e.g. fans to quickly vent the air out of the setup. See comments in the text. Please discuss the limitations of the experimental setup at the appropriate place in the discussion.

(5d) The stimulus air enters through a tube (what type of tube, diameter, length, etc) containing pressurized air (how was the air obtained into bags (type of bag, how is it sealed?), and the efflux directly into the flight chamber (how, nozzle?). However, it seems that there is no control of the efflux. How was leakage prevented, particularly how the bags were airtight sealed around the plants?

(5e) The plants were bagged in very narrowly fitting bags. The maize plants look bent and damaged, which probably explains the GLVs found in the samples. The Desmodium in the picture (Figure 5 supplement), which we should assume is at least a representative picture?) appears to be rather crammed into the bag with maize and looks in rather poor condition to start with (perhaps also indicating why they release these volatiles?). It would be good to describe the sampling of the plants in detail and explain that the way they were handled may have caused the release of GLVs.

(6) Figure 1 seems redundant as a main figure in the text. Much of the information is not pertinent to the paper. It can be used in a review on the topic. Or perhaps if the authors strongly wish to keep it, it could be placed in the supplemental material.

Reviewer #2 (Public review):

Based on the controversy of whether the Desmodium intercrop emits bioactive volatiles that repel the fall armyworm, the authors conducted this study to assess the effects of the volatiles from Desmodium plants in the push-pull system on behavior of FAW oviposition. This topic is interesting and the results are valuable for understanding the push-pull system for the management of FAW, the serious pest. The methodology used in this study is valid, leading to reliable results and conclusions. I just have a few concerns and suggestions for improvement of this paper:

(1) The volatiles emitted from D. incanum were analyzed and their effects on the oviposition behavior of FAW moth were confirmed. However, it would be better and useful to identify the specific compounds that are crucial for the success of the push-pull system.

(2) That would be good to add "symbols" of significance in Figure 4 (D).

(3) Figure A is difficult for readers to understand.

(4) It will be good to deeply discuss the functions of important volatile compounds identified here with comparison with results in previous studies in the discussion better.

Author Response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

The manuscript of Odermatt et al. investigates the volatiles released by two species of Desmodium plants and the response of herbivores to maize plants alone or in combination with these species. The results show that Desmodium releases volatiles in both the laboratory and the field. Maize grown in the laboratory also released volatiles, in a similar range. While female moths preferred to oviposit on maize, the authors found no evidence that Desmodium volatiles played a role in lowering attraction to or oviposition on maize.

Strengths:

The manuscript is a response to recently published papers that presented conflicting results with respect to whether Desmodium releases volatiles constitutively or in response to biotic stress, the level at which such volatiles are released, and the behavioral effect it has on the fall armyworm. These questions are relevant as Desmodium is used in a textbook example of pest-suppressive sustainable intercropping technology called push-pull, which has supported tens of thousands of smallholder farmers in suppressing moth pests in maize. A large number of research papers over more than two decades have implied that Desmodium suppresses herbivores in push-pull intercropping through the release of large amounts of volatiles that repel herbivores. This premise has been questioned in recent papers. Odermatt et al. thus contribute to this discussion by testing the role of odors in oviposition choice. The paper confirms that ovipositing FAW preferred maize, and also confirmed that odors released from Desmodium appeared not important in their bioassays.

The paper is a welcome addition to the literature and adds quality headspace analyses of Desmodium from the laboratory and the field. Furthermore, the authors, some of whom have since long contributed to developing push-pull, also find that Desmodium odors are not significant in their choice between maize plants. This advances our knowledge of the mechanisms through which push-pull suppresses herbivores, which is critically important to evolving the technique to fit different farming systems and translating this mechanism to fit with other crops and in other geographical areas.

Thank you for your careful assessment of our manuscript.

Weaknesses:

Below I outline the major concerns:

(1) Clear induction of the experimental plants, and lack of reflective discussion around this: from literature data and previous studies of maize and Desmodium, it is clear that the plants used in this study, particularly the Desmodium, were induced. Maize appeared to be primarily manually damaged, possibly due to sampling (release of GLV, but little to no terpenoids, which is indicative of mostly physical stress and damage, for example, one of the coauthor's own paper Tamiru et al. 2011), whereas Desmodium releases a blend of many compounds (many terpenoids indicative of herbivore induction). Erdei et al. also clearly show that under controlled conditions maize, silver leaf and green leaf Desmodium release volatiles in very low amounts. While the condition of the plants in Odermatt et al. may be reflective of situations in push-pull fields, the authors should elaborate on the above in the discussion (see comments) such that the readers understand that the plant's condition during the experiments. This is particularly important because it has been assumed that Desmodium releases typical herbivore-induced volatiles constitutively, which is not the case (see Erdei et al. 2024). This reflection is currently lacking in the manuscript.

We acknowledge the need for a more reflective discussion on the possible causes of volatile emission due to physical damage. Although the field plants were carefully handled, it is possible that some physical stress may have contributed to the release of volatiles, such as green leaf volatiles (GLVs). We ensured the revised manuscript reflects this nuanced interpretation (lines 282 – 286). However, we also explained more clearly that our aim was to capture the volatile emission of plants used by farmers under realistic conditions and moth responses to these plants, not to be able to attribute the volatile emission to a specific cause (lines 115 – 117). We revised relevant passages throughout the results and discussion to ensure that we do not make any claims about the reason for volatile emissions, and that our claims regarding these plants and their headspace being representative of the system as practiced by farmers are supported. In the revised manuscript we provide a new supplementary table S2 that additionally shows the classification of the identified substances, which also shows that the majority of the substances that were found in the headspace of the sampled plants of Desmodium intortum or Desmodium incanum are monoterpenes, sesquiterpenes, or aromatic compounds, and not GLVs (that are typically emitted following damage).

(2) Lack of controls that would have provided context to the data: The experiments lack important controls that would have helped in the interpretation:

2a The authors did not control the conditions of the plants. To understand the release of volatiles and their importance in the field, the authors should have included controlled herbivory in both maize and Desmodium. This would have placed the current volatile profiles in a herbivory context. Now the volatile measurements hang in midair, leading to discussions that are not well anchored (and should be rephrased thoroughly, see eg lines 183-188). It is well known that maize releases only very low levels of volatiles without abiotic and biotic stressors. However, this changes upon stress (GLVs by direct, physical damage and eg terpenoids upon herbivory, see above). Erdei et al. confirm this pattern in Desmodium. Not having these controls, means that the authors need to put the data in the context of what has been published (see above).

We appreciate this concern. Our study aimed to capture the real-world conditions of push-pull fields, where Desmodium and maize grow in natural environments without the direct induction of herbivory for experimental purposes (lines 115 – 117). We agree that in further studies it would be important to carry out experiments under different environmental conditions, including herbivore damage. However, this was not within the scope of the present study.

2b It would also have been better if the authors had sampled maize from the field while sampling Desmodium. Together with the above point (inclusion of herbivore-induced maize and Desmodium), the levels of volatile release by Desmodium would have been placed into context.

We acknowledge that sampling maize and other intercrop plants, such as edible legumes, alongside Desmodium in the push-pull field would have allowed us to make direct comparisons of the volatile profiles of different plants in the push-pull system under shared field conditions. Again, this should be done in future experiments but was beyond the scope of the present study. Due to the amount of samples we could handle given cost and workload, we chose to focus on Desmodium because there is much less literature on the volatile profiles of field-grown Desmodium than maize plants in the field: we are aware of one study attempting to measure field volatile profiles from Desmodium intortum (Erdei et al. 2024) and no study attempting this for Desmodium incanum. We pointed out this justification for our focus on Desmodium in the manuscript (lines 435 - 439). Additionally, we suggested in the discussion that future studies should measure volatile profiles from all plants commonly used in push-pull systems alongside Desmodium (lines 267 – 269).

2c To put the volatiles release in the context of push-pull, it would have been important to sample other plants which are frequently used as intercrop by smallholder farmers, but which are not considered effective as push crops, particularly edible legumes. Sampling the headspace of these plants, both 'clean' and herbivore-induced, would have provided a context to the volatiles that Desmodium (induced) releases in the field - one would expect unsuccessful push crops to not release any of these 'bioactive' volatiles (although 'bioactive' should be avoided) if these odors are responsible for the pest suppressive effect of Desmodium. Many edible intercrops have been tested to increase the adoption of push-pull technology but with little success.

We very much agree that such measurements are important for the longer-term research program in this field. But again, for the current study this would have exploded the size of the required experiment. Regarding bioactivity, we have been careful to use the phrase "potentially bioactive" solely when referring to findings from the literature (lines 99–103), in order to avoid making any definitive claims about our own results.

Because of the lack of the above, the conclusions the authors can draw from their data are weakened. The data are still valuable in the current discussion around push-pull, provided that a proper context is given in the discussion along the points above.

We think our revisions made the specific aims of this study more explicit and help to avoid misleading claims.

(3) 'Tendency' of the authors to accept the odor hypothesis (i.e. that Desmodium odors are responsible for repelling FAW and thereby reduce infestation in maize under push-pull management) in spite of their own data: The authors tested the effects of odor in oviposition choice, both in a cage assay and in a 'wind tunnel'. From the cage experiments, it is clear that FAW preferred maize over Desmodium, confirming other reports (including Erdei et al. 2024). However, when choosing between two maize plants, one of which was placed next to Desmodium to which FAW has no tactile (taste, structure, etc), FAW chose equally. Similarly in their wind tunnel setup (this term should not be used to describe the assay, see below), no preference was found either between maize odor in the presence or absence of Desmodium. This too confirms results obtained by Erdei et al. (but add an important element to it by using Desmodium plants that had been induced and released volatiles, contrary to Erdei et al. 2024). Even though no support was found for repellency by Desmodium odors, the authors in many instances in the manuscript (lines 30-33, 164-169, 202, 279, 284, 304-307, 311-312, 320) appear to elevate non-significant tendencies as being important. This is misleading readers into thinking that these interactions were significant and in fact confirming this in the discussion. The authors should stay true to their own data obtained when testing the hypothesis of whether odors play a role in the pest-suppressive effect of push-pull.

We appreciate this feedback and agree that we may have overstated claims that could not be supported by strict significance tests. However, we believe that non-significant tendencies can still provide valuable insights. In the revised version of the manuscript, we ensured a clear distinction between statistically significant findings and non-significant trends and remove any language that may imply stronger support for the odor hypothesis than what the data show in all the lines that were mentioned.

(4) Oviposition bioassay: with so many assays in close proximity, it is hard to certify that the experiments are independent. Please discuss this in the appropriate place in the discussion.

We have pointed this out in the submitted manuscript in lines 275 – 279. Furthermore, we included detailed captions to figure 4 - supporting figure 3 & figure 4 - supporting figure 4. We are aware that in all such experiments there is a danger of between-treatment interference, which we pointed out for our specific case. We stated that with our experimental setup we tried to minimize interference between treatments by spacing and temporal staggering. We would like to point out that this common caveat does not invalidate experimental designs when practicing replication and randomization. We assume that insects are able to select suitable oviposition sites in the background of such confounding factors under realistic conditions.

(5) The wind tunnel has a number of issues (besides being poorly detailed):

5a. The setup which the authors refer to as a 'wind tunnel' does not qualify as a wind tunnel. First, there is no directional flow: there are two flows entering the setup at opposite sides. Second, the flow is way too low for moths to orient in (in a wind tunnel wind should be presented as a directional cue. Only around 1.5 l/min enters the wind tunnel in a volume of 90 l approximately, which does not create any directional flow. Solution: change 'wind tunnel' throughout the text to a dual choice setup /assay.)

We agree with these criticisms and changed the terminology accordingly from ‘wind tunnel’ to ‘dual choice assay’. We have now conducted an additional experiment which we called ‘no-choice assay’ that provides conditions closer to a true wind tunnel. The setup of the added experiment features an odor entry point at only one side of the chamber to create a more directional airflow. Each treatment (maize alone, maize + D. intortum, maize + D. incanum, and a control with no plants) was tested separately, with only one treatment conducted per evening to avoid cross-contamination, as described in the methods section of the no-choice assay.

5b. There is no control over the flows in the flight section of the setup. It is very well possible that moths at the release point may only sense one of the 'options'. Please discuss this.

We added this to the discussion (lines 369 – 374). The new no-choice assays also address this concern by using a setup with laminar flow.

5c. Too low a flow (1,5 l per minute) implies a largely stagnant air, which means cross-contamination between experiments. An experiment takes 5 minutes, but it takes minimally 1.5 hours at these flows to replace the flight chamber air (but in reality much longer as the fresh air does not replace the old air, but mixes with it). The setup does not seem to be equipped with e.g. fans to quickly vent the air out of the setup. See comments in the text. Please discuss the limitations of the experimental setup at the appropriate place in the discussion.

We added these limitations to the discussion and addressed these concerns with new experiments (see answer 5a).

5d. The stimulus air enters through a tube (what type of tube, diameter, length, etc) containing pressurized air (how was the air obtained into bags (type of bag, how is it sealed?), and the efflux directly into the flight chamber (how, nozzle?). However, it seems that there is no control of the efflux. How was leakage prevented, particularly how the bags were airtight sealed around the plants?

We added the missing information to the methods and provided details about types of bags, manufacturers, and pre-treatments in the method section. In short, PTFE tubes connected bagged plants to the bioassay setup and air was pumped in at an overpressure, so leakage was not eliminated but contamination from ambient air was avoided.

5e. The plants were bagged in very narrowly fitting bags. The maize plants look bent and damaged, which probably explains the GLVs found in the samples. The Desmodium in the picture (Figure 5 supplement), which we should assume is at least a representative picture?) appears to be rather crammed into the bag with maize and looks in rather poor condition to start with (perhaps also indicating why they release these volatiles?). It would be good to describe the sampling of the plants in detail and explain that the way they were handled may have caused the release of GLVs.

We included a more detailed description of the plant handling and bagging processes to the methods to clarify how the plants were treated during the dual-choice and the no-choice assays reported in the revised manuscript. We politely disagree that the maize plants were damaged and the Desmodium plants not representative of those encountered in the field. The plants were grown in insect-proof screen houses to prevent damage by insects and carefully curved without damaging them to fit into the bag. The Desmodium plant pictured was D. incanum, which has sparser foliage and smaller leaves than D. intortum.

(6) Figure 1 seems redundant as a main figure in the text. Much of the information is not pertinent to the paper. It can be used in a review on the topic. Or perhaps if the authors strongly wish to keep it, it could be placed in the supplemental material.

We think that Figure 1 provides essential information about the push-pull system and the FAW. To our knowledge, this partly contradictory evidence so far has not been synthesized in the literature. We realize that such a figure would more commonly be provided in a review article, but we do not think that the small number of studies on this topic so far justify a stand-alone review. Instead, the introduction to our manuscript includes a brief review of these few studies, complemented by the visual summary provided in Figure 1 and a detailed supplementary table.

Reviewer #2 (Public review):

Based on the controversy of whether the Desmodium intercrop emits bioactive volatiles that repel the fall armyworm, the authors conducted this study to assess the effects of the volatiles from Desmodium plants in the push-pull system on behavior of FAW oviposition. This topic is interesting and the results are valuable for understanding the push-pull system for the management of FAW, the serious pest. The methodology used in this study is valid, leading to reliable results and conclusions. I just have a few concerns and suggestions for improvement of this paper:

(1) The volatiles emitted from D. incanum were analyzed and their effects on the oviposition behavior of FAW moth were confirmed. However, it would be better and useful to identify the specific compounds that are crucial for the success of the push-pull system.

We fully agree that identifying specific volatile compounds responsible for the push-pull effect would provide valuable insights into the underlying mechanisms of the system. However, the primary focus of this study was to address the still unresolved question whether Desmodium emits detectable or “significant” amounts of volatiles at all under field conditions, and the secondary aim was to test whether we could demonstrate a behavioral effect of Desmodium headspace on FAW moths. Before conducting our experiments, we carefully considered the option of using single volatile compounds and synthetic blends in bioassays. We decided against this because we judged that the contradictory evidence in the literature was not a sufficient basis for composing representative blends. Furthermore, we think it is an important first step to test f. or behavioral responses to the headspaces of real plants. We consider bioassays with pure compounds to be important for confirmation and more detailed investigation in future studies. There was also contradictory evidence in the literature regarding moth responses to plants. We thus opted to focus on experiments with whole plants to maintain ecological relevance.

(2) That would be good to add "symbols" of significance in Figure 4 (D).

We report the statistical significance of the parameters in Figure 4 (D) in Table 3, which shows the mixed model applied for oviposition bioassays. While testing significance between groups is a standard approach, we used a more robust model-based analysis to assess the effects of multiple factors simultaneously. We provided a cross-reference to Table 3 from the figure description of Figure 4 (D) for readers to easily find the statistical details.

(3) Figure A is difficult for readers to understand.

Unfortunately, it is not entirely clear which specific figure is being referred to as "Figure A" in this comment. We tried to keep our figures as clear as possible.

(4) It will be good to deeply discuss the functions of important volatile compounds identified here with comparison with results in previous studies in the discussion better.

Our study does not provide strong evidence that specific volatiles from Desmodium plants are important determinants of FAW oviposition or choice in the push-pull system. Therefore, we prefer to refrain from detailed discussions of the potential importance of individual compounds. However, in the revised version, we provide an additional table S2 which identifies the overlap with volatiles previously reported from Desmodium, as only the total numbers are summarized in the discussion of the submitted paper.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The points raised are largely self-explanatory as to what needs to be done to fully resolve them. At a minimum the text needs to be seriously revised to:

(1) reflect the data obtained.

(2) reflect on the limitations of their experimental setup and data obtained.

(3) put the data obtained and its limitations in what these tell us and particularly what not. Ideally, additional headspace measurements are taken, including from herbivory and 'clean' maize and Desmodium (in which there is better control of biotic and abiotic stress), as well as other crops commonly planted as companion crops with maize (but none of them reducing pest pressure).

Thank you for this summary. Please see our detailed responses above.

In addition to the main points of critique provided above, I have provided additional comments in the text (https://elife-rp.msubmit.net/elife-rp_files/2024/07/18/00134767/00/134767_0_attach_28_25795_convrt.pdf). These elaborate on the above points and include some new ones too. These are the major points of critique, which I hope the authors can address.

Thank you very much for these detailed comments.

Reviewer #2 (Recommendations for the authors):

It is important to note that the original push-pull system was developed against stemborers and involved Napier grass (still used) around the field, which attracts stemborer moths, and Molasses grass as the intercrop that repels the moths and attracts parasitoids. Later, Molasses grass was replaced by desmodiums because it is a legume that fixes nitrogen and therefore can increase nitrate levels in the soil, but most importantly because it prevents germination of the parasitic Striga weed. The possible repellent effect of desmodium on pests and attraction of natural enemies was never properly tested but assumed, probably to still be able to use the push-pull terminology. This "mistake" should be recognized here and in future publications. It is a real pity that the controversy over the repellent effect of desmodium distracts from the amazing success of the push-pull system, also against the fall armyworm.

We thank the reviewer for pointing out these issues, which are part of the reason for our Figure 1 and why we would like to keep it. We have described this development of the system in the introduction to better present the push-pull system. Our aim in Figure 1 and Table S1 is to highlight both the evidence of the system's success, and the gaps in our understanding, regarding specifically control of damage from the FAW.

Version 2

Abstract

Abstract

Push-pull systems for sustainable pest management combine repellent stimuli from intercrops (“push”) and attractive stimuli from border plants (“pull”) to repel herbivorous insects from a main crop and attract the herbivores’ natural enemies. The most widespread implementation, intercropping the legume Desmodium with maize surrounded by border grass, reduces damage from the invasive fall armyworm (FAW) Spodoptera frugiperda. However, the three publications to date investigating underlying mechanisms disagree about the role of the Desmodium intercrop and whether it emits volatiles that repel FAW. Here, we synthesize knowledge on the effects of maize-Desmodium push-pull on FAW, report volatiles from two commonly used Desmodium species: Desmodium intortum and D. incanum, and test whether these affect the behavior of gravid FAW moths in bioassays. We detected 25 volatiles from field-grown Desmodium, many in the headspaces of both species, including volatiles previously reported to repel lepidopteran herbivores. In cage oviposition assays, FAW moths preferred to oviposit on maize over Desmodium, but not on maize further from, versus closer to Desmodium plants that were inaccessible to the moths, but sharing the air. In flight tunnel assays, moths approached the headspace of maize more than the shared headspace of maize and Desmodium, but pairwise differences were often insignificant. Thus, headspaces of Desmodium species include volatiles that could repel FAW moths, and gravid moths were generally more attracted to maize and its headspace than to either Desmodium species or mixed maize-Desmodium headspaces. However, our results indicate that volatiles from Desmodium are not sufficient to explain reduced FAW infestation of maize under push-pull cultivation.

Introduction

The sustainable intensification of agriculture is essential to achieve Sustainable Development Goal (SDG) 2, zero hunger, and reduce hidden costs of meeting requirements of other SDGs such as 13, climate action (FAO et al., 2020). This is particularly important in the developing Global South where larger yield gaps are observed compared to countries with greater financial resources (Tilman et al., 2011). Intercropping may offer sustainable solutions with potential for strong positive effects on pest and disease control as well as associated biodiversity (Beillouin et al., 2021).

The concept of “push-pull” mixed cropping was first reported in 1987 as a “stimulo-deterrence” strategy for the reduction of pest damage in cotton, combining a pest repellent intercrop (“push”) interspersed with the main crop, and an attractant border crop (“pull”) to trap pests at the field perimeter (Bruce et al., 1987). As most widely practiced, current push-pull systems benefit from the attractive and repellent stimuli of perennial inter- and border crops (referred to as companion crops), thereby reducing or eliminating the need for pesticides (Pickett et al., 2014). In Kenya, the first push-pull system, introduced in 1997, was designed to reduce yield losses caused by stemborer species and comprised a combination of the main crop maize (Zea mays L), the intercrop molasses grass (Melinis minutiflora), and the border crop Napier grass (Pennisetum purpureum) or Sudan grass (Sorghum vulgare sudanense) (Khan et al., 1997b).

Several volatiles detected from molasses grass repelled female stem borers and simultaneously attracted their parasitoid Cotesia sesamiae (Khan et al., 1997a). Following the initial success of this system, there was interest in replacing molasses grass with a leguminous repellent intercrop that could improve soil quality by fixing nitrogen while providing high-quality fodder. The legumes silverleaf Desmodium (Desmodium uncinatum) and its congeneric (Desmodium intortum) were shown to repel ovipositing stem borers and furthermore to suppress parasitic Striga weeds. (Khan et al., 2000, 2002). The first Desmodium-intercropped version of push-pull used D. uncinatum in place of M. minutiflora together with P. purpureum. As a result of screening for more drought-tolerant companion plants, a version using D. intortum combined with the border crop Brachiaria cv Mulato II was developed as “climate-adapted push-pull”, which has been demonstrated to increase maize yield by a factor of 2.5 (Midega et al., 2015). Recently, a “third generation” push-pull system was evaluated using the intercrop D. incanum to replace D. intortum – which does not flower and set seed in many parts of tropical Africa and thus limits adoption and spread of push-pull by limiting seed supply for the intercrop – with Brachiaria brizantha cv Xaraes as border crop, which is more resistant to herbivorous mites that attack the mulato cultivar (Cheruiyot et al., 2021).

The invasive fall armyworm (FAW), Spodoptera frugiperda, originating from a maize-specialized strain from northern and central America, has become a major threat to African maize crops since 2016 (Goergen et al., 2016; Day et al., 2017; Hailu et al., 2021; Zhang et al., 2023). It spread rapidly through East Africa, with most of the farmers in Ethiopia and Kenya encountering the FAW after the long rains in the first half of 2017 (Kumela et al., 2018). The invasiveness of the moth is fueled by its relatively short life cycle of about four weeks and the capability of adult females to lay hundreds of eggs (Sparks, 1979). Other traits favoring the FAW’s spread in Africa and beyond are its strong flight capacity, lack of diapause, reported survival in diverse habitats, rapid development of resistance to insecticides and viruses, and polyphagous nature (Wan et al., 2021). Although the FAW has a wide host range of at least 353 species across 76 plant families (Montezano et al., 2018), the high financial and yield losses in maize due to this invasive FAW strain are particularly devastating (Eschen et al., 2021). The FAW is reported to outcompete resident stemborers (Hailu et al., 2021; Sokame et al., 2021; Mutyambai et al., 2022). The use of pesticides is a popular approach to control the FAW, posing risks to environmental and human health including acute pesticide-related illnesses, as many smallholder farmers do not use personal protective equipment while spraying (Tambo et al., 2020). Therefore, sole reliance on pesticides is not sufficient to manage FAW sustainably, and Integrated Pest Management (IPM) strategies, such as the promotion of natural enemies, are desirable (Nyamutukwa et al., 2022; Van den Berg and du Plessis, 2022). Climate-adapted and third-generation push-pull systems have been reported to reduce plant damage and yield loss caused by the FAW (Midega et al., 2018; Hailu et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021; Mutyambai et al., 2022).

A systematic review on the chemical ecology of push-pull systems by Lang et al. (2022) found only thirty publications (seven primary sources, and 23 publications reporting on results from these primary sources) on the chemistry of push-pull and related mixed-cropping systems, from which 206 substances were reported to be potentially associated with push and pull effects. Of these, 101 were categorized as plant volatiles and reported by studies with a focus on plant-insect interactions (as opposed to studies with a focus on Striga control). However, none of these publications reported volatiles sampled under field conditions. Two recent papers, both published after the literature review conducted by Lang et al. (2022), reported additional potentially bioactive substances from companion plants in maize-Desmodium-grass push-pull systems, and both publications supported the hypothesis that volatiles from the companion crop D. intortum repel the FAW and attract parasitic wasps (Sobhy et al., 2022; Peter et al., 2023). In contrast, a third recent publication by Erdei et al. (2024) detected low levels of volatiles from D. intortum plants and found that the presence of D. intortum upwind from maize did not reduce oviposition on maize, while FAW larvae offered D. intortum fed on it and became trapped by its silicate trichomes; the authors thus proposed that reduced FAW infestation in maize-Desmodium intercropping settings is more likely to result from physical trapping of FAW larvae by Desmodium.

The protection mechanisms of push-pull systems are assumed to be complex and to operate across multiple interactions, protecting maize not only from FAW but also from other pests and parasitic weeds, such as Striga. However, further research is needed to understand the mechanisms by which the maize-Desmodium-grass push-pull system can protect maize from FAW. A summary of the studies to date on push-pull effects on the FAW can be found in Figure 1 and supplementary table S1. To our knowledge, no data on plant volatiles from the more drought-resistant D. incanum have been published so far. Therefore, this paper focuses on the volatiles of D. intortum and D. incanum, aiming to capture the headspace of plants used by farmers under realistic conditions, rather than attributing the composition to specific causes such as damage. Additionally, our goal is to assess the effects of exposure to the headspaces of these plants, versus direct exposure to the plants, on the interaction of FAW with maize. We collected volatiles from plants in farmers’ fields as well as in bioassay setups under semi-field conditions, and conducted oviposition and flight tunnel bioassays to assess FAW moth preferences.

Studies to date on push-pull effects on the fall armyworm (FAW) embedded in the mechanisms according to the current state of knowledge of the system.

Volatiles (and potentially other traits) of the intercrop repel the herbivorous insect and additionally attract its parasitoids, while volatiles (and potentially other traits) of the border crop attract herbivores away from the main crop (Pickett et al., 2014; Khan et al., 2018; Eigenbrode et al., 2016). The pull effect of the border crop Napier grass, Pennisetum purpureum was observed in earlier systems, as it attracted stemborers (Khan et al., 1997b), but could not be confirmed for the FAW with the border crop Brachiaria cv Mulato II (Sobhy et al., 2022). For more detail on all studies, see supplementary table S1.

Results

Desmodium Volatile Profiles

A total of 25 substances were measured in at least 2/3 of field-collected samples from each of the Desmodium species, of which 11 occurred in both species. These substances include (E)-β-ocimene, (Z)-3-hexen-1-ol, (Z)-3-hexen-1-ol, acetate, 1-octen-3-ol, 3-octanone, caryophyllene, (3E)-4,8-dimethyl-1,3,7-nonatriene (DMNT), germacrene d, indole, linalool and (3E,7E)-4,8,12-Trimethyl-1,3,7,11-tridecatetraene (TMTT). Five peaks were not fully identifiable and were categorized based on mass spectra and relative retention times as a benzene derivative, a naphthalene derivative, a monoterpene, and two sequiterpenes (for feature information, see Table 1). The sesquiterpene (E)-α-bergamotene was detected only in the headspace of D. incanum, while another unidentified sesquiterpene only occurred in D. intortum (see Figure 2 for a heatmap with means, Figure 3 for sample comparison, and Figure 2—figure Supplement 1 for a per-sample heatmap). Of these substances, (Z)-1,5-octadien-3-ol, 1,2,3-trimethylbenzene, 3-pentanol, and hexyl acetate are here for the first time reported in connection with the chemical ecology of push-pull cropping systems (for more detailed information of each volatile substance, see supplementary table S2). Germacrene D and 1-hexanol were not yet reported from Desmodium species, but were reported in relation to mixed cropping for pest control (Lang et al., 2022). Germacrene D was detected in elevated levels in maize plants grown in soil from push-pull fields in comparison with those grown in soil from maize monocultures (Mutyambai et al., 2019), while germacrene D and 1-hexanol both were reported to be emitted from bean plants, Vicia faba, and used by the black bean aphid, Aphis fabae, for host detection (Pickett and Khan, 2016; Webster et al., 2008).

Heatmap comparing the log10-transformed peak area of the non-zero hits of substances present in at least 2/3 of the samples of at least one of the Desmodium species in the field, showing relative peak areas detected in field- or pot-grown (bioassay conditions) D. intortum and D. incanum, with pot-grown maize for comparison.

Peak intensity can only be meaningfully compared within the same substance. The x-axis displays the species and their growth conditions (field or pot) and the y-axis shows the target substances in alphabetical order. The color indicates mean of the log-10 transformed peak area of all samples with the substance present. The grey color indicates that a substance occurs in more than one, but less than 2/3 of the samples from the field. “Maize Infested” refers to maize plants that were exposed to FAW eggs, as moths were allowed to oviposit on the plants two nights prior to volatile sampling. Sample sizes: Ambient Control Field = 4, D. incanum Field = 14, D. incanum Pot = 3, D. intortum Field = 11, D. intortum Pot = 5, Maize infested Pot = 4, Maize Pot = 4. Figure 2—figure supplement 1. Per-sample heatmap

A: Principal component analysis plot based on the normalized peak areas of the 25 target substances. Sample sizes: Ambient Control Field = 4, D. incanum Field = 14, D. incanum Pot = 3, D. intortum Field = 11, D. intortum Pot = 5, Maize infested Pot = 4, Maize Pot = 4 B: Loading plot of the projection of the variables on the first two dimensions. The x-axis displays PC1 (Dimension 1) and the y-axis shows PC2 (Dimension 2). The color indicates the cos2, whereby dark and long arrows indicate a better representation of a loading on these first two dimensions.

Oviposition Choice Bioassays

Oviposition choice bioassays were conducted in cages to determine the influence of Desmodium plants and their headspace on the oviposition behavior of FAW moths. A fraction of the potted Desmodium plants used in the oviposition assays was sampled for volatiles the night before the start of trials, and the composition of these samples was compared with the substances and features detected from Desmodium plants sampled in farmers’ fields. In the comparison between field Desmodium plants, and potted bioassay plants that were sampled in a smaller number, the majority of substances showed a higher relative abundance in potted D. incanum (bioassay conditions), whereas in D. intortum, some substances detected in field samples occurred in higher concentrations in the potted plants, while others were not detected. Sixteen of the volatiles detected in Desmodium plants in farmers’ fields were also present in either uninfested or infested potted maize plants (see Figure 2). In direct contact treatments, FAW moths were allowed to oviposit directly on Desmodium plants within the cage, and in treatments labelled as ‘indirect’, moths could chose between two maize plants, whereby one was closer to a Desmodium plant placed immediately outside the cage and thus closer to the headspace of that Desmodium plant (see Figure 4). Except for these ‘indirect’ treatments, plants were separated by at least 50 cm within and between cages, and positions were randomized (see methods and materials).

A: Setup of the control with two maize plants inside the cage. B: Setup of the treatments with direct contact with one maize and one Desmodium plant inside the cage. C: Setup of the treatments with indirect contact with two maize plants placed inside the cage and one Desmodium plant placed in proximity to one of the maize plants, but unreachable for the moths. D: The boxplots display the relative number of eggs (light grey boxes) or egg batches (dark grey boxes) per position (x axis) and treatment (superordinate boxes). The lower and upper hinges correspond to the 25th and 75th percentiles, while the whiskers extend to the largest and smallest non-outlier values, respectively. Outliers are values outside a window of 1.5 x the interquartile range and are plotted individually. Desmodium is abbreviated with ‘Des’ and treatment is abbreviated with ‘treat’. See Table 3 for a model-based statistical analysis and Figure 4—figure Supplement 1 for a breakdown of eggs per repetition and batch. Sample sizes (n, replicate units of a given treatment): Control = 21, D. incanum direct = 19, D. intortum direct = 20, D. incanum indirect = 19, D. intortum indirect = 20. Figure 4—figure supplement 1. Total egg count per repetition and per batch during oviposition bioassays Figure 4—figure supplement 2. Schemes and photographs of the experimental setup of the choice oviposition bioassays Figure 4—figure supplement 3. Scheme of the cage setup in greenhouse 1B Figure 4—figure supplement 4. Scheme of the cage setup in greenhouse 3B Figure 4—figure supplement 5. Photographs of the egg collection process

The count of eggs and egg batches showed that fewer eggs were laid directly on Desmodium plants, with the egg count on maize plants being 7.9 or 6.8 times higher than on D. incanum or D. intortum, respectively, corresponding to a similar difference in the number of egg batches (Figure 4 and Figure 4—figure Supplement 1). A mixed model with egg counts on the maize plant as the observed variable and accounting for plant position detected a significant difference between the control (two maize plants, no neighboring Desmodium) and all other treatments (F(1) = 11.11, p = 0.0012), as well as for the comparison of direct versus indirect contact with Desmodium plants (F(1) = 9.51, p = 0.0026). However, no significant difference was found between the two Desmodium species (F(1) = 1.08, p = 0.3012) (see Table 3).

Dual-Choice Assays

Dual-choice assays were conducted to compare short-term effects of plant volatiles on the flight behavior of FAW moths. The setup permitted only low air flow relative to its total volume and so the possible mixing of stimuli both across trials and within a trial is not well defined and is likely more complex than in most laboratory trials, and more similar to field and other less controlled settings. Overpressure was used to ensure the simultaneous transfer of headspaces from both treatments at equal airflow rates and from opposite directions into the setup, and moth movement toward these opposing sources was interpreted as choice. See the description of the methods and materials as well as the no-choice setup, which addresses limitations of this choice setup.

Two-thirds of the moths tested showed ten or fewer segment changes during the five minutes of each experiment. Across all treatments, at least 77% of the moths showed no movement in the last 2 min of the experiment or only little activity, with a maximum of two segment changes in the last 3 min of the experiment, which was interpreted to mean that they had made a decision within the five-minute time frame of the experiment. Overall, moths showed a reduction in segment changes over time, particularly for the comparison of maize vs. maize + D. intortum (see Figure 5). Across all treatments, a few individuals displayed high flight activity of up to 76 segment changes with no indication of decline over time.

A: Line plot displaying the number of segment changes per minute until the experiment ended after 5 min. The line width indicates the number of overlaying lines and the colors indicate the moth activity level: ‘Settled after 2 min’ = No segment changes after 2 min. ‘Settled after 3 min’ = No segment changes after 3 min. ‘Little Activity’ = Moths showed max. 2 segment changes in the last 3 min of the experiment. ‘High Activity’ = Moths showed higher activity with at least 6 segment changes in the last 3 min of the experiment. The mean of all repetitions is represented by the black dashed line. B: Setup of the dual-choice assay for a Desmodium treatment comparing a maize plant (left) vs. a maize and a Desmodium plant (right). The moths are released through an opening in the center (at 0 cm) and observed for 5 min. C: Violin plot displaying the duration of stays in each segment in treatments with D. incanum and D. intortum (comparison of maize and maize + Desmodium) with the inclusion of the individual data points per repetition (black) and their mean (red): The x-axis displays the location in the dual-choice assay and the y-axis the total duration that each moth stayed in each segment. Below the x-axis, the last position of the moths that were considered settled (settled after <3 min or little activity) is displayed. See Table 4 for the statistical analysis and Figure 5—figure Supplement 1 for the set-up and segmentation of the dual-choice assay setup. Sample sizes (n, replicate units of a given treatment: Control = 18, D. incanum = 18, D. intortum = 19.) Figure 5—figure supplement 1. Photograph of the experimental setup of the dual-choice assay

No significant effects were detected when comparing the stays in the maize segments versus the Desmodium segments across the treatments using a mixed model (see Table 4). However, when comparing maize alone to maize with D. intortum, FAW moths tended to prefer the D. intortum treatment, with 2.5 times more individuals settling in that area and spending, on average, twice as long there. In the comparison between maize and maize with D. incanum, only minor differences were observed in the mean residence time (less than 1 minute); however, twice as many moths settled on the D. incanum side (see Figure 5). Fourteen of nineteen moths exposed to airflow from maize vs. maize + D.intortum settled in a segment of the dual choice assay within the first three minutes of the experiment, a larger proportion than in the maize vs. maize or the maize vs. maize + D. incanum treatment.

No-Choice Assays

No-choice assays were conducted to assess the short-term effects of plant volatiles on the flight behavior of FAW moths in a setup with higher airflow and conducted in a blocked manner so that no mixing could occur between treatments (see methods and materials). Treatments included a no-plant control as well as maize or maize + Desmodium for both Desmodium species. Repetitions in which moths exhibited movement in the final 2 minutes of the trial were considered non-decisive and excluded.

The initial distance, representing the first movement bigger than 5 cm made by a moth, did not show significant differences between treatments. However, the highest mean initial distance was observed in the maize treatment, while the maize + D. intortum treatment exhibited the lowest mean. The landing distance, defined as the final position relative to the release point, showed more pronounced treatment effects. Specifically, the maize treatment had the highest mean landing distance, while the maize + D. incanum treatment had the lowest (see Figure 6). Statistical analysis revealed a significant difference between the maize and the Desmodium treatments (see Table 5).

A: Setup of the no-choice assay for a Desmodium treatment. The moths are released through an opening in the right side (at 0 cm) and their movement observed for 5 min. The plant odor was released from the outlet at 100 cm, while the moths were allowed to move freely up to 115 cm. All repetitions that showed activity in the last two minutes were excluded, as they are considered to have made no clear decision. B: Violin plot with the inclusion of the individual data points per repetition displaying the distance flown in the first, initial movement and the final location that was considered the finally chosen position. The x-axis displays the treatment and the y-axis the total distance that a moth moved or flew. See Table 5 for the statistical analysis. The sample sizes corresponded to 19 - 20 for all samples, whereby all repetitions with activity in the last 2 min of the experiment were removed. Remaining sample sizes (n, replicate units of a given treatment): Control = 17, Maize = 17, Maize + D. incanum = 16, Maize + D. intortum = 17.

Discussion

The fall armyworm (FAW) Spodoptera frugiperda invaded sub-Saharan Africa in 2016 and is responsible for major yield losses from maize crops (Day et al., 2017; Goergen et al., 2016). Push-pull cropping of maize with Desmodium and border grasses is reported to reduce FAW damage (Midega et al., 2015; Cheruiyot et al., 2021; Mutyambai et al., 2022; Yeboah et al., 2021; Hailu et al., 2018) and to increase maize yield (Midega et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021). Push-pull cropping systems benefit from the attractive and repellent stimuli of perennial companion crops (Pickett et al., 2014), with more than one hundred plant volatiles from different classes, especially terpenoids and fatty acid-derived “green leaf” volatiles, reported to be associated with push-pull effects (Lang et al., 2022). Nevertheless, the repellent role of volatiles from greenleaf Desmodium (D. intortum) has been investigated in only three studies, with conflicting reports of whether these volatiles repel FAW moths, and the only study reporting field measurements recovering few volatiles in relatively low abundance (Sobhy et al., 2022; Peter et al., 2023; Erdei et al., 2024).

We characterized volatiles sampled from both D. intortum and a recently introduced alternative Desmodium intercrop, D. incanum, growing in farmer fields or in bioassay conditions, and conducted bioassays with FAW moths to determine whether the presence or proximity of Desmodium plants or their headspace would alter moth preferences for maize. We identified up to 17 - 19 substances in headspace samples from Desmodium plants, often in abundance similar to or greater than volatiles measured from maize plants under comparable conditions. We observed that FAW moths preferred to oviposit on maize rather than on Desmodium plants of either species, but that close proximity to a Desmodium plant sharing the airspace did not influence FAW oviposition preferences for maize in a greenhouse cage choice assay. Consistent with this, in no-choice assays, FAW flying upwind landed closer to a maize odor source than when maize was combined with either species of Desmodium, but in dual-choice assays, FAW moths did not stay significantly closer to maize headspace sources over headspace sources from a combination of maize and Desmodium plants. We note that the choice assays (oviposition and dual-choice movement) were performed in relatively complex odor backgrounds, whereas the no-choice assay was performed with a simpler odor environment.

Furthermore, all assays were performed in a risk area for malaria and other mosquito-borne illnesses, so experimenters were protected with DEET, which has low volatility but is active in the vapor phase, where it can cause host avoidance by binding to insect olfactory receptors (although this has been mostly studied in the dipterans Aedes aegypti and Drosophila melanogaster) (DeGennaro, 2015). We reduced any potential influence on the lepidopteran in our experiments by using minimal amounts of DEET, avoiding direct contact of any DEET with experimental materials, and conducting behavioral assays without the insects being exposed to volatiles from the experimenter (experimenters were generally not present during oviposition assays, and only small amounts of DEET were used during the movement assays indoors in Nairobi).

We conclude that Desmodium plants in push-pull cultivation emit volatiles that have the potential to repel FAW (and to attract their natural enemies), but our bioassays provide only weak support for the hypothesis that these volatiles repel FAW from maize. While (perhaps, subtler) effects of the Desmodium headspace might be more reliably detected using more refined experimental setups, our mixed support of this hypothesis is in line with the reports in literature thus far (Figure 1). We suggest that other features of the maize-Desmodium-border grass push-pull system, including the influence of border grasses as well as belowground interactions and other traits of Desmodium, are important for protecting maize from FAW moths and their larvae. We furthermore call for field or semi-field bioassays that better capture the choice of moths exposed to push-pull versus non-push-pull fields, in contrast to individual headspace environments presented out of context, or maize plants closer to vs. further from Desmodium plants in pots.

Emission of Volatiles Implicated in Push-Pull by Desmodium Intercrops

The headspace of Desmodium plants was dynamically sampled on Tenax TA adsorbent directly in push-pull fields, as well as from potted plants used for oviposition bioassays, overnight. Due to limited resources and time, we chose to focus on the volatile profiles of field-grown Desmodium rather than field-grown maize or intercropped edible legumes, as this aligned best with our aim of assessing the role of Desmodium volatiles. Additionally, we opted not to use individual volatile substances in bioassays because the conflicting evidence in the literature made it difficult to select meaningful blends, and testing plant emissions first was crucial for ensuring ecological relevance in our experiments. However, we recommend that future studies sample the volatiles of all plants commonly used in newly developed push-pull systems, including the edible legumes that are now more often intercropped, allowing for a contextualized comparison of volatile profiles.

We excluded all substances found in fewer than two-thirds of the field samples from further analysis. While this threshold is arbitrary and may overlook potential differences in plant stressors across fields, we chose to focus on substances commonly present in field-grown Desmodium plants. Quantitative comparisons within single substances are possible, but must be handled with care, as normalization based on the biomass and leaf area of the plants was not possible. The chosen adsorbent, Tenax TA, is known to be effective for lipophilic to medium-polarity organic compounds of intermediate molecular weight (approximately C7-C26) (Dettmer and Engewald, 2002; Tholl et al., 2020). It is possible that headspace samples changed during storage (up to 3 months tightly closed at room temperature) as degradation or thermal rearrangements on Tenax TA material have been reported (Dettmer and Engewald, 2002; Alborn et al., 2021). We extracted headspace samples using thermal desorption, which is simple and sensitive (Tholl et al., 2006), but not appropriate for some substances, such as the sesquiterpene germacrene A, which are susceptible to thermal rearrangement or degradation (Faraldos et al., 2007). While the potted plants were kept in insect-proof greenhouses, it cannot be excluded that field plants might have been exposed to insect herbivory, though representative and healthy plants were selected for the volatile sampling. Although plants were handled carefully, it is still possible that volatile release could have resulted from physical damage.

Despite these caveats, our samples yielded a variety of volatiles with high signal:noise ratios, of which 13 had been previously reported from sampling the headspace of potted D. intortum plants onto Porapak Q for sampling durations of 24 h or 48 h followed by solvent extraction and GC-MS analysis (Sobhy et al., 2022; Peter et al., 2023). In contrast, Erdei et al. (2024) detected fewer volatiles in lower relative abundance from D. intortum using solid-phase microextraction (SPME) and a saturation time of 18 - 24h to sample intact, mechanically damaged or herbivory-induced potted plants or intact plants in Tanzanian and Ugandan fields. Our study is the only one of these that limited the window of sampling strictly to the nighttime hours corresponding to the reported activity window for the FAW (Sparks, 1979). In pretests, no shorter activity window could be determined as the moths still showed mating behaviour and potential oviposition at around midnight (see Appendix 1). Differences may be explained in part by different sampling techniques, as the sensitivity of SPME can be lower in comparison to dynamic headspace collection (Tholl et al., 2006). To our knowledge, the volatile profile of D. incanum was not previously reported. In conclusion, our data are consistent with the hypothesis that D. intortum and D. incanum emit volatiles, many of which have been previously associated with push-pull effects, the repellence of lepidopteran herbivores or the attraction of their natural enemies.

Oviposition Choice Bioassays

FAW oviposition behaviour was observed over three nights in two treatments per Desmodium species. In one treatment, FAW moths had direct contact with Desmodium plants and in another, they had only indirect contact through aerial exposure to plants outside of, but directly next to a mesh cage. Moths always had direct contact with maize plants.

Critical consideration must be given to differences in the volatile profiles between the potted plants and those measured in farmers’ fields, which led to a separation of potted and field D. intortum in the second dimension of a PCA, and a near-separation of D. incanum in the first dimension. As the untargeted analysis was conducted based on the field Desmodium samples, any volatile substances that only occur in the potted plants were not identified or analyzed further (although full results of the GC-MS analysis are available with the source data for this publication). Oviposition on maize can alter the maize volatile profile and affect volatile emission in the following days, which is why the maize samples are divided into plants before (control) and after (infested) oviposition exposure. Of the substances occurring in field Desmodium plants, approximately half were abundant in healthy and infested potted maize plants, with only one substance, (E)-α-Bergamotene, being present solely in infested maize plants. The sampling of potted plants was conducted using the same procedure as in the field, but in a greenhouse with netted walls on two sides that allowed air circulation, as used for the oviposition bioassays. Overall, this highlights the importance of studying oviposition cues and moth preferences directly in push-pull fields in the future.

In all treatments, the number of eggs laid per batch showed no difference depending on the position, which indicates that FAW moths did not adjust their behavior once they started laying eggs. This differs slightly from the outcomes reported by Peter et al. (2023), where a reduction in egg number per batch was observed and the number of eggs, but not the number of egg batches, differed between maize vs. maize + D. intortum. However, a clear preference for maize compared to the two Desmodium species was observed in the treatments where FAW moths had direct contact with both plant species, with a large effect size of 7 – 8 times more eggs (as measured by number of eggs or number of batches) on maize than on Desmodium. The preference of maize over D. intortum in oviposition assays with direct contact to both plants is consistent with various publications using similar setups (Sobhy et al., 2022; Peter et al., 2023; Erdei et al., 2024). There was little oviposition on the cage in our bioassays, which indicates that moths primarily chose between plants in our setup. This differs from the strong preference found by Sobhy et al. (2022) of FAW moths to oviposit directly on the mesh of a bioassay cage rather than on the plants when offered maize co-planted with D. intortum. Desmodium and maize plants were planted in the same pots by Sobhy et al. (2022), and therefore maize volatiles may have been affected by belowground interactions or through priming by Desmodium volatiles prior to or during oviposition bioassays. This is in fact closer to the configuration in push-pull fields, but does not isolate the influence of the Desmodium headspace, as we sought to do here.

In our bioassays, no significant effect could be seen in the treatments where moths had only indirect contact with Desmodium volatiles coming from a Desmodium plant proximate to one maize plant, but outside of the cage. This is inconsistent with the hypothesis that Desmodium volatiles repel FAW moths as proposed (Pickett et al., 2014; Sobhy et al., 2022). It is possible that the distance of about 60 cm between the plants in the cages was not suitable to create a gradient of volatiles sufficient to allow moths to choose between maize plants closer to versus further from Desmodium, in which case neighboring treatments with distances of 50 – 100 cm could have also affected each other. Although the distance reflected the approximate spacing between Desmodium and maize plants in push-pull fields, this limitation must be considered, as the close proximity of assays may have compromised the ability to create distinct volatile environments. To our knowledge, the only comparable setup was reported by Erdei et al. (2024), in which two maize plants were exposed to volatiles of either D. intortum or an artificial plant in a modified dual-choice assay, and moths did not show a significant preference for either maize plant.

In conclusion, we confirm a preference of FAW moths to oviposit on maize rather than on D. intortum or D. incanum when directly offered plants. However, no significant influence of the Desmodium headspace was observed. It is not to be expected that herbivores use all potential host plants equally, and preferences among host plants or for host over non-host plants can be based on different plant traits. Ovipositing FAW moths may also prefer maize over Desmodium due to other traits such as the surface texture, as proposed in the findings from Erdei et al. (2024) that silicate trichomes on D. intortum pose a danger to FAW larvae; or chemical repellents directly on the leaf surface. Belowground interactions or priming between Desmodium and maize plants co-planted in push-pull fields may also change moths’ preference for maize, an effect which could not have been observed with our experimental design, but might explain inconsistencies with findings from Sobhy et al. (2022).

Dual-Choice and No-Choice Assays

The short-term flight behavior of FAW moths was tested in a dual-choice assay setup where headspaces of a maize plant vs. maize + Desmodium were compared against each other, and a no-choice assay with a single headspace source upwind of the moth. Flight behavior showed large variation among the individual moths, but three-quarters of the moths had settled after at least three minutes or showed few changes in the last two minutes of each experiment. Though flow rates were measured at the incoming tubing of the dual-choice setup and the exhaust was placed above the center (where the moth was also introduced), this experimental setup allowed no control over the exact flows in the flight arena and the individual flows were low compared to the total arena volume. Therefore, the exact mixing of the opposing treatments is unknown, and cross-contamination between treatments was also possible. Furthermore, moths may behave differently over longer times or when given more space to maneuver, as under normal field conditions. The quarter of the moths that showed high activity indicates the difficulty of quantifying the responses of such active fliers. Interestingly, a larger proportion of moths settled after 2 minutes when presented with maize versus maize + D. intortum in the dual-choice assay, rather than continuing to fly. While some tendencies were apparent in this setup, none of the treatments showed any significant preference. In contrast, a significant repellent effect of Desmodium added to maize, versus maize alone, was observed in the no-choice assay as judged by final landing position. The combination of maize + D. incanum showed a greater repellent effect than maize + D. intortum. Similar repellence of moths by D. intortum headspace with or without the addition of maize was observed in no-choice assays reported in the literature when the total distance flown upwind at the end of the experiment was used as a metric (Sobhy et al., 2022; Peter et al., 2023).

Conclusion

Several field studies have described positive effects of push-pull systems with the companion crops D. intortum or D. incanum on maize yield (Midega et al., 2015, 2018; Cheruiyot et al., 2021). Here, we show that 17 - 19 volatile substances, including plant volatiles previously indicated to repel lepidopteran hervivores or attract their natural enemies, were found in the headspaces of both companion crops D. intortum and D. incanum within the temporal activity window of FAW moths. However, we did not observe a repellent effect of Desmodium plant headspaces on FAW moths in most bioassays. FAW moths clearly preferred to oviposit on maize over D. intortum or D. incanum plants. Our results thus indicate that moths prefer maize over Desmodium and that this may be in-fluenced by short-range mechanisms such as an unfavorable leaf surface of Desmodium. However, the main interest in the context of push-pull is how Desmodium contributes to protecting co-planted maize. A reduction of FAW damage on maize under push-pull cultivation has been demonstrated in multiple studies (Midega et al., 2018; Hailu et al., 2018; Cheruiyot et al., 2021; Yeboah et al., 2021; Mutyambai et al., 2022). Based on the literature, effects of border crop volatiles may be important for reducing FAW damage in push-pull fields, in combination with the attraction of parasitoids by Desmodium volatiles and unfavorable host qualities of Desmodium plants for FAW (Figure 1). Furthermore, the Desmodium intercrop may be equally or more important for weed suppression, soil fertility, and improving maize plant health and rigor as for volatile-mediated repellence of FAW in the push-pull system. For example, recent studies showed that planting maize in soil collected from push-pull fields or co-planting it with silverleaf Desmodium uncinatum changed the composition of volatiles and defensive non-volatile substances in maize plants (Mutyambai et al., 2024; Bass et al., 2024). Future studies wishing to test the importance of Desmodium volatiles in the system could best resolve this question by adding and subtracting Desmodium headspace, or manipulating moths’ direct access to Desmodium plants, in experimental push-pull and control fields, thus ensuring relevant context while avoiding experimental artifacts.

Methods and Materials

All volatile sampling and bioassays were performed with two Desmodium species, Desmodium intortum (greenleaf Desmodium) and Desmodium incanum (often referred to as African Desmodium by the push-pull farmers in Western Kenya), referred to in the methods section collectively as Desmodium for simplicity. All raw data and code used for the statistical analyses can be found on Zenodo (CERN, Geneva, Switzerland, https://doi.org/10.5281/zenodo.11633890) and GitHub (https://github.com/DariaMOdermatt/DesmodiumVolatilesinPush-Pull).

Pretest Moth Activity Window

The determination of the activity window of the FAW was relevant for the choice of the duration of the volatile sampling. The FAW is reported to be nocturnal (Sparks, 1979), without reports of a more precise activity window. For a first cycle, two single maize leaves of the landrace ‘Jowi white’ were placed in wooden cages with meshed side walls (100 x 60 x 60 cm) in high cylindrical glasses filled with water, but sealed with a piece of cotton to prevent moths from entering the glasses. Moths (3 - 4 days old) were released in pairs of one female and one male in 10 different cages (resulting in 20 moths in total) and observed from 7 pm - 11.30 pm over two consecutive nights. The moths were re-collected in falcon tubes in the morning after the first night, sealed with a piece of cotton to allow air circulation and stored at a temperature of approx. 25 °C until they were released in the same combinations on the following evening. For the second night, fresh maize leaves were provided for the moths to oviposit. For a second cycle, potted maize plants with multiple leaves of the landrace ‘Jowi white’ were placed inside the wooden cages with a piece of cotton soaked with water. Three female and two male moths of different ages were placed each in three cages (resulting in 15 moths in total) and observed from 6 pm - 1 am in two consecutive nights. During daytime the moths and the plant were left unaltered in the cage.

Headspace Sampling

We chose to focus on the volatile profiles of field-grown Desmodium rather than field-grown maize or intercropped edible legumes. This decision was driven by the relative scarcity of literature on the volatile profiles of Desmodium plants in real-world conditions of push-pull fields. To our knowledge, only one study has attempted to measure the volatiles of D. intortum Erdei et al. (2024), and no studies have measured volatiles of D. incanum in fields. Headspace samples of Desmodium were collected from plants in push-pull fields of local smallholder farmers near Mbita in the counties Homa bay and Siaya, coordinates determined with WGS84 coordinate system (latitude, longitude): D. incanum = (−0.382096, 34.175487), (−0.4298279, 34.207891), (0.189957, 34.36072), D. intortum = (−0.382096, 34.175487), (−0.6302736, 34.494595), (−0.551479, 34.314673). Desmodium headspace samples were collected in three different push-pull fields in May and June 2023 during the long rainy season, selecting four representative and healthy plants per field. As is becoming more common, some of the push-pull fields were mixed with vegetables such as kale or cowpeas. Headspace collections were also performed from potted Desmodium plants, as well as from healthy and infested potted maize plants on the research campus. Five three-week-old maize plants, later labeled as “infested”, were exposed in two cages, each with each three female and two male moths, for two nights prior to headspace sampling, which resulted in 1 to 10 batches per plant. Potted plants were placed in a greenhouse with a glass roof and two glass walls, in which air was able to circulate through netted walls on the opposing longer walls of the greenhouse.

As an adsorbent, commercially available Tenax TA glass tubes (containing 150 mg adsorbent, 35 – 60 mesh, Markes International Ltd, England or 186 mg adsorbent, 60 – 80 mesh, Merck, Germany) were used. A handful of leaves were enclosed in a roasting bag (Sainsbury’s, London, UK), which was heated to 200 °C for at least 1 hour to clean the bag, then cooled to room temperature. The inlet tubing pushed 500 mL/min charcoal-filtered air into the bags from the lower bag rim, while the Tenax TA tube was connected to an air outlet tubing through a small hole at the top of the bag. Airflows were regulated by PYE Volatile Collection Kits (B. J. Pye, Kings Walden, UK) as described in Steen et al. (2019). Volatiles were accumulated on the adsorbent by aspirating the air at 200 mL/min (+/- 40 mL/min) for 11.75 h (+/- 30 min) overnight from 7 pm until 6.45 am on the following day. Under the same conditions, four ambient controls were sampled with an empty roasting bag on two fields with each Desmodium species. Additionally, a Tenax TA storage control was stored with all tubes, but never taken to the fields, and analyzed to check for contamination.

TD-GC-MS Measurement

All samples were measured on a TD-GC-MS instrument (Thermal Desorption – Gas Chromatography – Mass Spectrometry) from Shimadzu (TD30-QP2020NX, Kyoto, Japan). The Tenax TA tubes were desorbed for 15 min at 220 °C under 80 mL/min nitrogen flow. Volatiles were trapped on a Tenax cooling trap at −20 °C and after desorption was completed, the cooling trap was heated rapidly to 230 °C. The sample was injected with a split ratio of 1:5 and separated on a Rtx-Wax column (30 m, 0.25 mmID, 0.25 μm, Restek Corporation, USA) with the following 35 min oven program: holding at 40 °C for 2 min, heating to 150 °C with 10 °C/min, holding at 150 °C for 2 min, heating to 190 °C with 3 °C/min, heating to 230 °C with 10 °C/min and holding at 230 °C for 3 min. Mass fragments from 33 m/z to 400 m/z resulting from electron ionization at 70 eV were recorded with a scan rate of 5 Hz at an ion source temperature of 230 °C from 3 – 30 min.

Feature Detection

A feature list was generated in MZmine (Version 3.9.0, (Schmid et al., 2023)) from all substances that occurred in at least one-third of the field samples of either Desmodium species and in less than 4/5 of the combined control samples (ambient controls and storage control). Features were numbered and considered as unidentified if there was no corresponding reference substance. After removal of contaminants, the features were added to an existing target list that was composed with commercially available plant volatiles that can be mainly classified as terpenoids and benzenoids, which as a whole contain a large variety of functional groups such as alcohols, aldehydes, ketones and esters. Additionally, the existing list included four green leaf volatiles ((E)-2-hexen-1-ol acetate, (E)-2-hexenal, (Z)-3-hexen-1-ol, (Z)-3-hexen-1-ol acetate) and a selection of more specific substances reported in push-pull fields such as the norsesquiterpene (E)-4,8–dimethyl–1,3,7-nonatriene (DMNT) and the norditerpene (E,E)-4,8,12-trimethyl-1,3,7,11-tridecatetraene (TMTT). The peak integration for all substances and features was manually checked for all control samples (ambient and storage control) using the LabSolutions Insight GCMS software (Shimadzu corporation, Kyoto, Japan). Peaks for all targets that occurred in at least 1/2 of the Desmodium field samples and in no more than 3/5 of controls were also manually checked for all samples. Finally, all targets that occurred in at least 2/3 of the field samples of D. intortum or D. incanum and in no more than 3/5 of the controls were considered as present. In Table 1 the electron impact ionization (EI)-MS data of the present, unidentified features are displayed, and in Table 2 the origin of the reference standards used to identify the target hits can be found.

Oviposition Choice Bioassays

Plants

All plants were planted in plastic pots in black cotton soil in insect-proof greenhouses in Mbita, Kenya. Maize plants (SC Duma 43, Seed Co Limited, Nairobi, Kenya) were grown from seeds with the addition of fertilizer (3 g DAP 18-46-0, Yara East Africa LTD, Nairobi, Kenya) and used at the age of 2 – 4 weeks with five to seven fully-grown leaves. Desmodium plants were obtained from a push-pull field on the icipe Mbita campus and kept in pots without fertilizer.

Fall Armyworm Moths

To form the FAW colony, wild individuals were collected in the counties Siaya, Kisumu, Migori, and Vihiga in Western Kenya. Larvae were fed on artificial diet based on soy flour, wheat germ, raw linseed oil, mineral mix, sugar, aureomycin, vitamins, agar, methyl parabene, sorbic acid, and calcium propionate (Article No 870-265-3747, Southland Products Inc, Lake Village AR, USA). Moths were fed on water and in rare cases after emergence on 10% honey solution. The colony was occasionally restocked with wild FAW moths. Each day after hatching, moths of both sexes were transferred to a cage to allow mating until the start of the experiment.

Experimental Procedure

Cages (100 x 60 x 60 cm) lined with wooden floors and ceilings and netted walls were placed in greenhouses that allowed air circulation through mesh side walls (for pictures of the real setup, see Figure 4—figure Supplement 2, for exact positioning in the greenhouses, see Figure 4—figure Supplement 3 and Figure 4—figure Supplement 4). Two plants were placed within the cage at the greatest distance possible (approx. 60 cm) in two opposite corners. For the control treatment, two maize plants were placed in the cage, while the direct Desmodium treatment consisted of one maize plant and one Desmodium plant in the cage. In two additional treatments with indirect contact, two maize plants were placed in the cage, while one Desmodium plant was placed outside the cage unreachable for the FAW moths, but in proximity to one maize plant. Three female and two male moths aged 4 – 5 days were released approx. 1 h before dusk and allowed to oviposit under natural conditions of L12:D12 for three consecutive nights. Eggs were collected in five different groups depending on the position in which they were laid: either on one of the two plants, close to one of the plants within a maximum of 20 cm distance, or on the cage further than 20 cm from any plant (labeled as “No Decision”). Each treatment was repeated 19 – 21 times in total over five cycles, with up to five replicates per treatment and cycle.

Egg Count

Each batch of FAW eggs was collected with sticky tape to separate the layers and spread all eggs out to one dimension. Each egg batch was taped to white paper and photographed in a UV imaging system (Syngene, Cambridge, England) against UV light (312 nm) coming from underneath the paper. A script for semi-automatic counts in ImageJ (Version 1.54f, National Institutes of Health, USA) was developed and used for counting the exact number of eggs per picture (see Figure 4— figure Supplement 5). The code is available in the data and code repository.

Dual-Choice and No-Choice Assays

Plants

All plants were planted from seeds without the addition of fertilizer in plastic pots in standard red soil mixed with manure (ratio of 2:1) at the icipe campus in insect-proof greenhouses in Nairobi, Kenya. Maize plants (SC Duma 43, Seed Co Limited, Nairobi, Kenya) were used at the age of 3 – 7 weeks with five to eight fully-grown leaves. It was observed that maize plants seemed to grow more slowly in Nairobi than in Mbita, which could be explained by the generally lower temperatures in Nairobi.The seeds of D. intortum and D. incanum were collected from a push-pull field located on the icipe campus in Mbita. These seeds were subsequently propagated through the replanting of cuttings, which were periodically trimmed to ensure healthy growth.

Fall Armyworm Moths

To form the FAW colony, wild individuals were collected in Central Kenya such as in the counties Kiambu, Muranga, and Embu and restocked with wild individuals every three months. Larvae were reared on maize leaves, while moths were only fed with water. The colony was occasionally restocked with wild FAW moths. Each day after hatching, moths of both sexes were transferred to a cage to allow mating until the start of the experiment. At least 2 h prior to the experiment, female moths were separated from males and placed in the darkened experimental room for adjustment to temperature and humidity.

Dual-Choice Assay: Experimental Procedure

The five upper leaves of a maize plant were carefully wrapped in a preheated (200 °C for at least 1 h) and cooled roasting bag (Sainsbury’s, London, UK) with the addition of approx. the same biomass of Desmodium on one side of a dual-choice assay setup (100 x 30 x 30 cm) and compared to a single maize plant wrapped in the same manner on the other side. Maize leaves were carefully curved without damaging them to fit into the bag, while the bag was secured to the bottom edge with little tension using plastic straps supplied with the oven bags. Control treatments were conducted with a single maize plant wrapped in this way on each side. Air was pushed with the help of a Volatile Collection Kit (B. J. Pye, Kings Walden, UK) through activated charcoal filters and into the roasting bag from the lower rim of the bag at a rate of 740 – 820 mL/min with a PTFE tube with a diameter of approx. 2 mm. A PTFE tube with a diameter of approx. 10 mm and a length of approx. 120 cm conveyed the plant volatiles from the top of the roasting bag to one side of the dual-choice assay. Inside the dual-choice assay, the air was pumped out from the center at the 0 cm mark resulting in an air stream at the air transfer tubing of both sides of the dual-choice assay setup. Airflow was measured each day on both sides using a portable flow meter (Vögtlin Instruments GmbH, Switzerland) and ranged between 480 and 570 mL/min across different days, with a maximum difference of 4 mL/min between the two sides. The higher incoming airflow created over-pressure in the bags, which prevented unfiltered air from entering the dual-choice system. One female moth at an age of 4 – 5 days was released through a hole in the center at 0 cm and her flight behavior was observed for 5 minutes. The dual-choice assay was separated into 5 segments of 20 cm each, and a time stamp was set every time the moth changed segments (See Figure 5—figure Supplement 1). All experiments were conducted between 7 pm and 1:30 am under red light. Each trial was recorded on video using a mobile phone camera (Fairphone 4, 48 MP), which captured the dual-choice assay from a front view. The videos were re-watched for data acquisition and in cases where moths were barely visible, complemented with notes from the live observation. To simplify the data analysis, the two left segments (50 – 30 cm left & 30 – 10 cm left) were combined, as were the two right segments (10 – 30 cm right, 30 – 50 cm right). An equal number of repetitions for each treatment was conducted consecutively each night, with the order of treatments randomized. Changing each set of plants took on average 15 minutes, while the room was ventilated for at least 30 minutes with ambient air between treatment changes.

No-Choice Assay: Experimental Procedure

The five upper leaves of a maize plant were carefully wrapped in a preheated (100 °C for at least 1 h) and cooled roasting bag (Sealapack, Manchester, UK) with the addition of approx. the same biomass of Desmodium on one side of a no-choice assay setup (115 x 33 x 33 cm). Maize leaves were carefully curved without damaging them to fit into the bag, while the bag was secured to the bottom edge with little tension using plastic straps supplied with the oven bags. Maize treatments were conducted with one maize plant only, and control treatments were performed with an empty roasting bag. Air was drawn through activated charcoal filters and directed into the roasting bag via a PTFE tube (approx. 2mm in diameter) at a flow rate of 1.5 L/min, with the air entering the bag from its lower rim. This process was facilitated using a Volatile Collection Kit (B. J. Pye, Kings Walden, UK). A PTFE tube was used, with a diameter of approx. 2 mm and a length of 120 cm, to transport plant volatiles from the top of the roasting bag to the source on the side of the no-choice assay. On the opposite side, air was evacuated at the 0 cm mark, creating a uniform air stream that flowed across the entire chamber of the no-choice assay 30 cm/s. The higher incoming airflow created over-pressure in the bags, which prevented unfiltered air from entering the dual-choice system. One female moth aged 4 – 5 days was released through a hole in the center at 0 cm and its flight behavior was observed for 5 minutes. The initial and final landing distances were noted, while the last movement was determined by videos that were recorded for each repetition. In cases where moths were not visible or only partially visible, we recorded their last clearly observed position as the final location, assuming any significant movement would have been captured on video. All experiments were carried out between 9 pm and 1 am under red light conditions. The same set of plants was used for five repetitions, with a different moth individual introduced for each repetition. Treatments were carried out on separate evenings to avoid any cross-contamination or carryover effects. All repetitions of a single treatment were conducted on the same evening, except for the maize (alone) treatment, which was carried out across two evenings.

Statistical analysis

All statistical analysis was performed in RStudio (R version 4.5.0, RStudio version 2024.12.1). Mixed models were used to determine significant effects of FAW egg positions in the oviposition bioassays, as well as the behavior of FAW moths in the dual-choice and no-choice assays. In the oviposition and dual-choice bioassays, control treatments with two maize plants were included. Therefore, the control replications were included twice in the dataset, once with the egg count data or the duration of stay of both sides interchanged. For the oviposition bioassay, the egg counts on the maize or the Desmodium plant were compared amongst all treatments, with the inclusion of the greenhouse, the cage (combination of the cage group and the cage number), the start date, and the replication number as random effects. For the dual-choice assay, the duration of stay closer to the maize or the Desmodium plants were compared amongst all treatments, with the inclusion of the date and the replication number as random effects. For the no-choice assay, the initial flight distance and the landing flight distance was compared amongst all treatment, with the inclusion of the date and the replication as random effects.